---
title: "LFQ modelling results"
author: "FGCZ"
date: "`r format(Sys.time(), '%d %B, %Y')`"
output: 
  html_document:
    includes:
      in_header: fgcz_header.html
      after_body: fgcz_footer.html
params:
  pars: NULL
bibliography: bibliography.bib
editor_options: 
  chunk_output_type: console
---



```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = FALSE)
library(flextable)
pars <- params$pars


makefunction <- function(outpath) {
  outpath <- outpath
  res <-  function(..., collapse = NULL){
    x <- c(...)
    res <- paste(x, collapse = "")
    if (file.exists(file.path(outpath,res))) {
      return(res)
    }else{
      stop("File ", file.path(outpath,res) , " does not exist!")
    }
  }
  return(res)
}

paste_file <- makefunction(outpath = pars$outpath)

```


- project Id : `r pars$project_Id`
- order Id : `r pars$order_Id`
- parent work-unit Id : `r pars$workunit_Id`
- modelling implemented by : `r pars$author`
- input file : `r pars$inputMQfile`


# Method [^0]


Protein fold changes were computed based on peptide intensity values reported in the MaxQuant generated `peptides.txt` file, using linear mixed-effects models [@lmer2015].
We preprocess the peptide intensities reported in the peptides.txt file as follows: we removed all proteins with only one identified peptide, we remove intensities equal zero, non zero intensities are $\log_2$ transformed and modified using robust z-score transformation (using the median and median average deviation). For each protein, we fitted a mixed-effects model to the peptide intensities. Fold changes and p-values for each contrast (see table below) are computed based on this model [@lmerTest2017]. For a contrast, all the protein p-values are adjusted for multiple testing using the Benjamini and Hochberg procedure to obtain the false discovery rate (FDR) [see @BenjaminiHochberg1995].

To estimate fold-changes for proteins for which we could not fit a mixed-effects model because of an excess in missing measurements, we proceed as follows: First, we compute the mean intensity for all peptides for each condition. For the proteins with no measurements in that condition, we impute the peptide intensities using the mean of the $10\%$ smallest average peptide intensities computed in step one. Afterwards, we compute contrasts (differences between conditions) for each peptide and finally, the median of the peptide estimates is used to provide a per protein fold change estimate.  

[^0]: This section can be used in the material and methods section alongside the model formula and the contrast specifications given below.

# Data preprocessing and diagnostic plots

A set of diagnostic plots is generated from protein intensities. The diagnostic plots (PCA, heatmaps) are generated from transformed, normalized protein intensities [^1].


[^1]: The peptide intensities are $\log_2$ transformed and z-scored. The protein intensity are determined using the [Tukey Median polish](https://en.wikipedia.org/wiki/Median_polish) from the transformed and scored peptide intensities.  



The raw files are annotated with the following factors and factor levels:

```{r annotation}
table_facade(pars$annotation, "Sample annotations.")

```


`r if(file.exists("qc_results/retention_time_plot.png")){' - [Peptide intensity vs RT plot](qc_results/retention_time_plot.png){target="_blank"}'}`
- PCA plot ([pdf](`r paste_file("qc_results/ms_intensities_PCA_.pdf")`){target="_blank"},[html](qc_results/ms_intensities_PCA_.html){target="_blank"}) showing the samples according to the first and second principal component.
- [Heatmap based on protein intensity correlation among samples](`r paste_file("qc_results/ms_intensities_heatmap_correlation_.pdf")`){target="_blank"}.
- [Protein intensities heatmap](`r paste_file("qc_results/ms_intensities_heatmap_.pdf")`){target="_blank"}. The columns represent samples while the rows represent proteins. The protein intensities were z scored.
- The protein intensities in [excel xlsx format](`r paste_file("qc_results/ms_intensities_wide_.xlsx")`){target="_blank"}.
- A [pdf file](`r paste_file("qc_results/ms_intensities_inference_figures.pdf")`){target="_blank"} visualizing all the peptide and protein intensities (the peptide intensities are $\log_2$ transformed and [z-scored](https://en.wikipedia.org/wiki/Standard_score) ). 
- A quality control report of estimated protein intensities [html](`r paste_file("qc_results/ms_intensities_qc.html")`){target="_blank"}.


# Mixed Effects Model

```{r alttext, include=FALSE}
mixed_txt <- "The following mixed-effects model was fitted [see @lmer2015]: "
fixed_txt <- "The following linear model was fitted: "
is_mixed_model <- LFQService:::.is_mixed_model

```


`r if(is_mixed_model(pars$model)) {mixed_txt} else {fixed_txt}`

__`r format(formula(pars$model))`__

Using this model the following contrasts were computed. 

```{r contrasts}
Contrasts <- data.frame(contrast = names(pars$Contrasts), explicit = as.character(pars$Contrasts))
table_facade(Contrasts, "Contrasts (vs - versus, gv - given).")

```


The p values of the ANOVA analysis and p values of the contrasts were estimated using the R package lmerTest [see @lmerTest2017].


```{r filepaths, include=FALSE}
anovaXLSX <- file.path(pars$modelling_dir, "ANOVA_Model")
anovaPDF <- file.path(pars$modelling_dir, "ANOVA_p.values_Model.pdf")
contrastXLSX <- file.path(pars$modelling_dir, "foldchange_estimates")
histogramFC <- file.path(pars$modelling_dir, "contrasts_Histogram_FC_estimate_Model")
histogramPVal <-   file.path(pars$modelling_dir, "contrasts_Histogram_p.value_Model")
histogramPValAdj <- file.path(pars$modelling_dir, "contrasts_Histogram_p.value.adjusted_Model")
volcanoPVal <- file.path(pars$modelling_dir, "contrasts_Volcano_p.value_Model")
volcanoAdjModPval <- file.path(pars$modelling_dir, "contrasts_Volcano_moderated.p.value.adjusted_Model")
volcanoPValAdj <- file.path(pars$modelling_dir, "contrasts_Volcano_p.value.adjusted_Model")
maPlot <- file.path(pars$modelling_dir, "contrasts_MA_FC_estimate_Model")
```


```{r alttext2, include =FALSE}
t1 <- "- __Volcano plot of p-Values vs fold change estimate__ ([pdf]("
t2 <- paste_file(volcanoPVal,".pdf")
t3 <- "){target='_blank'}; [html]("
t4 <- paste_file(volcanoPVal,".html")
t5 <- "){target='_blank'}) which on the vertical axis shows the protein $-\\log_{10}(p~values)$ versus the the log2 fold changes on the horizontal axis."
volcanoPValue <- paste(c(t1,t2,t3,t4,t5), collapse = "")

if (!is_mixed_model(pars$model)) {
  t1 <- "- __Volcano plot of adjusted moderated p-Values vs fold change estimate__ ([pdf]("
  t2 <- paste_file(volcanoAdjModPval,".pdf")
  t3 <- "){target='_blank'}; [html]("
  t4 <- paste_file(volcanoAdjModPval,".html")
  t5 <- "){target='_blank'}) which on the vertical axis shows the protein $-\\log_{10}(p~values)$ versus the the log2 fold changes on the horizontal axis."
  volcanoModeratedAdjustedPValue <- paste(c(t1,t2,t3,t4,t5), collapse = "")
}
```



- For all contrasts and proteins we provide:
    - A __table__ with all fold change estimates and p values ([excel xlsx format](`r paste_file(contrastXLSX,".xlsx")`); [html](`r paste_file(contrastXLSX,".html")`){target="_blank"}) 
    for all proteins and contrasts. For proteins not quantified in one of the conditions, _pseudo_ estimates are provided [^2]. The columns `estimate` and `pseudo_estimate` hold the fold changes. The column `p.value.adjusted` holds the False Discovery Rate. 
    - __Histogram of fold changes__ ([pdf](`r paste_file(histogramFC, ".pdf")`){target="_blank"}) shows the distribution of the contrasts for each comparison.
    - __Histogram of p-Values__ ([pdf](`r paste_file(histogramPVal,".pdf")`){target="_blank"}) shows the distribution of the p-Values for each contrast. This plot is used to asses the quality of your data and of the model (see [How to interpret p-value histogram]( http://varianceexplained.org/statistics/interpreting-pvalue-histogram/)).
    - __Histogram of adjusted p-Values__ ([pdf](`r paste_file(histogramPValAdj,".pdf")`){target="_blank"}) shows the distribution of the false discovery rates (FDR - p values adjusted using the Benjamin and Hochberg method [see @BenjaminiHochberg1995]).
    `r if(FALSE){volcanoPValue}`
    - __Volcano plot of adjusted p-Values vs fold change estimate__ ([pdf](`r paste_file(volcanoPValAdj,".pdf")`){target="_blank"}; [html](`r paste_file(volcanoPValAdj, ".html")`){target="_blank"})  which on the vertical axis shows the $-\log_{10}(FDR)$ versus the the _log2_ fold changes on the horizontal axis.
    `r if(!is_mixed_model(pars$model)){volcanoModeratedAdjustedPValue}`
    - [__MA plot__](https://en.wikipedia.org/wiki/MA_plot) ([pdf](`r paste_file(maPlot,".pdf")`){target="_blank"}; [html](`r paste_file(maPlot, ".html")`){target="_blank"}). The vertical axis shows the contrast estimate (_log2 FC_) while the horizontal axis shows the mean of the measurments (this is also known as mean difference plot MD).
    - ANOVA results ([excel xlsx format](`r paste_file(anovaXLSX,".xlsx")`);[html](`r paste_file(anovaXLSX,".html")`){target="_blank"} and histogram of p-Values [pdf](`r paste_file(anovaPDF)`){target="_blank"}) which show which factors included in the model are significant.


[^2]:  To estimate a fold-changes for proteins were a mixed-effects model can not be fitted, because there is no measurement for on of the conditions, we proceed as follows: First, we compute the mean intensity of all peptides for that condition. For the proteins with no measurements, we impute the peptide intensities using the mean of the $10\%$ smallest average peptide intensities computed in step one. 
Afterwards, we compute contrasts (differences between conditions) for each peptide and finally, the median of the peptide estimates is used to provide a per protein fold change estimate.  
