---
title: "LFQ modelling results"
author: "FGCZ"
date: "`r format(Sys.time(), '%d %B, %Y')`"
output: 
  html_document:
    includes:
      in_header: fgcz_header.html
      after_body: fgcz_footer.html
params:
  pars: NULL
---



```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = FALSE)
pars <- params$pars
```


- project ID : `r pars$project_id`
- order ID : `r pars$order_id`
- parent work-unit name : `r pars$workunit_name`
- input file : `r pars$inputMQfile`
- modelling implemented by : `r pars$author`

The raw files are annotated with the following factors and factor levels:

```{r annotation}
knitr::kable(pars$annotation, caption="Annotation")
```

The following model was fitted: 

```{r echo TRUE}
pars$model
```

Using this model the following contrasts were computed.


```{r contrasts}
Contrasts <- data.frame(contrast = names(pars$Contrasts), explicit = pars$Contrasts)
knitr::kable(Contrasts, caption = "Contrasts (vs - versus, gv - given)")
```


- ANOVA results in [excel xlsx format](modelling_results/Anova_Model.xlsx) and [histogram of p-Values](modelling_results/Anova_p.values_Model.pdf) which shows which factors are most significant.
- For all contrasts and proteins:
    - A table with all fold change estimates and p-Values [excel xlsx format](modelling_results/foldchange_estimates.xlsx) with p-values for all proteins and contrasts. For proteins not quantified in one of the conditions, _pseudo_ estimates are provided.
    - [Histogram of p-Values](modelling_results/Contrasts_Histogram_p.value_Model.pdf) shows the distribution of the p-Values for each contrast to see which contrast produced the largest number of significant changes
    - [Histogram of adjusted p-Values](modelling_results/Contrasts_Histogram_p.value.adjusted_Model.pdf) shows the distribution of the p-Values after multiplicity adjustment for each contrast to see which contrast produced the largest number of significant changes.
    - [Volcano plot of p-Values vs fold change estimate](modelling_results/Contrasts_Volcano_p.value_Model.pdf) which on the vertical axis shows the protein p-Values versus the the log2 fold changes on the horizontal axis. 
    - [Volcano plot of adjusted p-Values vs fold change estimate](modelling_results/Contrasts_Volcano_p.value_Model.pdf)  which on the vertical axis shows the adjusted protein p-Values versus the the log2 fold changes on the horizontal axis.



In addition you can find descriptive statistics visualizations


- [PCA plot](qc_results/protein_intensities_PCA_.pdf) showing the samples according to the first and second principal component.
- [Heatmap based on sample correlation](qc_results/protein_intensities_heatmap_correlation_.pdf).
- [Protein Intensity heatmap](qc_results/protein_intensities_heatmap_.pdf).
- The protein intensities in [excel xlsx format](qc_results/protein_intensities_wide_.xlsx).
- A quality control report of estimated protein intensities [html](qc_results/protein_intensities_qc.html).




