---
title: "Benchmarking the proDA package using the Ionstar Dataset"
author: "FGCZ - (Draft)"
date: "`r format(Sys.time(), '%d %B, %Y')`"
output: 
  html_document:
    toc: true
    toc_depth: 3
    toc_float:
      toc_collapsed: true
papersize: a4
geometry: margin=.5in
vignette: >
  %\VignetteIndexEntry{Benchmarking the proDA package using the Ionstar Dataset} 
  %\VignetteEncoding{UTF-8}
  %\VignetteEngine{knitr::rmarkdown}
editor_options:
  chunk_output_type: console
---

```{r setup, include=FALSE} 
knitr::opts_chunk$set(echo = TRUE, warning = FALSE, message = FALSE)
evalAll <- require(proDA)
fromPEP <- TRUE

```


# proDA benchmark based on peptide.txt intensities and median polish


```{r fromPeptideTXT , eval = fromPEP}

datadir <- file.path(find.package("prolfquaData") , "quantdata")
inputMQfile <-  file.path(datadir,
                          "MAXQuant_IonStar2018_PXD003881.zip")
inputAnnotation <- file.path(datadir, "annotation_Ionstar2018_PXD003881.xlsx")
mqdata <- list()
mqdata$data <- tidyMQ_Peptides(inputMQfile)
length(unique(mqdata$data$proteins))
mqdata$config <- prolfqua::create_config_MQ_peptide()

annotation <- readxl::read_xlsx(inputAnnotation)
res <- add_annotation(
  mqdata$data,
  annotation,
  fileName = "raw.file"
)

mqdata$config$table$factors[["dilution."]] = "sample"
mqdata$config$table$factors[["run_Id"]] = "run_ID"
mqdata$config$table$factorDepth <- 1
mqdata$data <- setup_analysis(res, mqdata$config)


lfqdata <- LFQData$new(mqdata$data, mqdata$config)
lfqdata$data <- lfqdata$data %>% filter(!grepl("^REV__|^CON__", protein_Id)) 
lfqdata$filter_proteins_by_peptide_count()
sr <- lfqdata$get_Summariser()
sr$hierarchy_counts()

lfqdata$remove_small_intensities()
sr <- lfqdata$get_Summariser()
sr$hierarchy_counts()

tr <- lfqdata$get_Transformer()
subset_h <- lfqdata$get_copy()
subset_h$data <- subset_h$data %>% dplyr::filter(grepl("HUMAN", protein_Id))
subset_h <- subset_h$get_Transformer()$log2()$lfq
lfqdataNormalized <- tr$log2()$robscale_subset(lfqsubset = subset_h)$lfq

lfqAggMedpol <- lfqdataNormalized$get_Aggregator()
lfqAggMedpol$medpolish()
lfqtrans <- lfqAggMedpol$lfq_agg

```

To use `proDA`, we need to create an `SummarizedExperiment`. We use the `to_wide` function of `prolfqua` to get the data in in the `SummarizedExperiment` compatible format.


```{r createSummarizedExperiment, eval=evalAll}
library(SummarizedExperiment)
wide <- lfqtrans$to_wide(as.matrix = TRUE)
library(SummarizedExperiment)
ann <- data.frame(wide$annotation)
rownames(ann) <- wide$annotation$sampleName

se <- SummarizedExperiment(SimpleList(LFQ = wide$data), colData = ann)
se

```


## Defining Contrasts and computing group comparisons

As usual, two steps are required, first fit the models, then comptue the contrasts.


```{r runProda, eval=evalAll, include=TRUE}
# Not Run
fit <- proDA(se, design = ~ dilution. - 1,data_is_log_transformed = TRUE)
contr <- list()
contr[["dilution_(9/7.5)_1.2"]] <- data.frame(contrast = "dilution_(9/7.5)_1.2",
                                              test_diff(fit, contrast = "dilution.e - dilution.d"))
contr[["dilution_(7.5/6)_1.25"]] <- data.frame(contrast = "dilution_(7.5/6)_1.25",
                                               test_diff(fit, contrast = "dilution.d - dilution.c"))
contr[["dilution_(6/4.5)_1.3(3)"]] <- data.frame(contrast = "dilution_(6/4.5)_1.3(3)", 
                                                 test_diff(fit, contrast ="dilution.c - dilution.b"))
contr[["dilution_(4.5/3)_1.5"]] <- data.frame(contrast = "dilution_(4.5/3)_1.5", 
                                              test_diff(fit, contrast ="dilution.b - dilution.a" ))

bb <- bind_rows(contr)
# End Not Run

```



## Benchmarking

```{r setUPBenchmark, eval = evalAll}

bb$name %>% unique() %>% length()
ttd <- prolfqua::ionstar_bench_preprocess( bb , idcol = "name" )

benchmark_proDA <- make_benchmark(ttd$data,
                                   contrast = "contrast",
                                toscale = c("pval"),
                                fcestimate = "diff",
                                benchmark = list(
                                  list(sc = "diff", desc = TRUE),
                                  list(sc = "t_statistic", desc = TRUE),
                                  list(sc = "scaled.pval", desc = TRUE)
                                ),  
                                model_description = "proDA_medpolishInt",
                                model_name = "proDA_medpolishInt",
                                FDRvsFDP = list(list(sc = "adj_pval", desc = FALSE))
, hierarchy = c("name"), summarizeNA = "t_statistic"
)

sum(benchmark_proDA$smc$summary$name)
sumarry <- benchmark_proDA$smc$summary
table_facade(sumarry, caption = "nr of proteins with 0, 1, 2, 3 missing contrasts.")

```


```{r prepBenchmarkforComparison, include= FALSE, eval = evalAll}
xdd <- ttd$data %>% dplyr::rename(protein_Id = name ,
                                  contrast = contrast,
                                  estimate = diff,
                                  statistic = t_statistic,
                                  p.value = pval,
                                  FDR = adj_pval  
)

benchmark2_proDA <- make_benchmark(xdd, model_description = "proDA_medpolishInt", model_name = "proDA")
saveRDS(benchmark2_proDA, file = "../inst/Benchresults/benchmark2_lfqInt_proDA.RDS")

```

```{r rocCurve, fig.cap="ROC curves", eval = evalAll}
res <- benchmark_proDA$pAUC_summaries()
res$barp
```

```{r pAUC02, fig.cap="plot ROC curves", eval = evalAll}
#res$ftable
benchmark_proDA$plot_ROC(xlim = 0.2)
```

```{r fdrfdp, fig.cap = "plot FDR vs FDP",eval = evalAll}
benchmark_proDA$plot_FDRvsFDP()
```

```{r fdptpr,eval = evalAll}
benchmark_proDA$plot_FDPvsTPR()
```
