---
title: "Benchmarking normalization, aggregation and models using the Ionstar Dataset"
author: "FGCZ - (Draft)"
date: "`r format(Sys.time(), '%d %B, %Y')`"
output: 
  html_document:
    toc: true
    toc_depth: 3
    toc_float:
      toc_collapsed: true
papersize: a4
geometry: margin=.5in
vignette: >
  %\VignetteIndexEntry{Benchmarking normalization, aggregation and models using the Ionstar Dataset} 
  %\VignetteEncoding{UTF-8}
  %\VignetteEngine{knitr::rmarkdown}
editor_options: 
  chunk_output_type: console
---



Please download and install the `prolfquaData` package from github


```{r setup, include=FALSE}
knitr::opts_chunk$set(
echo = TRUE,
message = FALSE,
warning = FALSE,
fig.width = 5,
fig.height = 5
)
```


```{r loadRequiredPackages}
rm(list = ls())
library(conflicted)
library(prolfqua)
library(tidyverse)
library(dplyr)
conflicted::conflict_prefer("filter", "dplyr")

```

Decide if you work with all data or for speedup with subset of data:

```{r}
SUBSET <- FALSE
```


We start by loading the IonStar dataset and the annotation from the `prolfquaData` package. The method `add_annotation` adds the annotation to the data.

```{r loadMQData}

datadir <- file.path(find.package("prolfquaData") , "quantdata")
inputMQfile <-  file.path(datadir,
                          "MAXQuant_IonStar2018_PXD003881.zip")
inputAnnotation <- file.path(datadir, "annotation_Ionstar2018_PXD003881.xlsx")
mqdata <- tidyMQ_Peptides_Config(inputMQfile)
mean(grepl("CON_", mqdata$data$top_protein))

annotation <- readxl::read_xlsx(inputAnnotation)


res <- add_annotation(
  mqdata$data,
  annotation,
  fileName = "raw.file"
)

```

The `setup_analysis` asserts that all columns specified in the configruation are present in the data. For more details about the proflqua configuration see the vignette "Creating Configurations".


```{r annotateData}
mqdata$config$table$factors[["dilution."]] = "sample"
mqdata$config$table$factors[["run_Id"]] = "run_ID"
mqdata$config$table$factorDepth <- 1
mqdata$data <- setup_analysis(res, mqdata$config)

```



```{r filterProteinsByPeptideCount}
datafilt <- prolfqua::filter_proteins_by_peptide_count(mqdata$data, mqdata$config)$data
length(unique(datafilt$protein_Id))

```


## Data normalization


We will normalize the data using the 'LFQTransformer' class. Since we know that the Human proteins are the Matrix in the experiment we will normalize the data using HUMAN proteins only.
To this task we subset the dataset by filtering for HUMAN proteins only and then use the 'log2_robscale_subset' method.


```{r filterForHumans}

lfqdata <- LFQData$new(datafilt, mqdata$config)
tr <- lfqdata$get_Transformer()

subset_h <- lfqdata$get_copy()
subset_h$data <- subset_h$data %>% filter(grepl("HUMAN", protein_Id))
lfqdataNormalized <- tr$log2_robscale_subset(lfqsubset = subset_h)

```


The figures below show the intensity distribution before and after normalization.

```{r plotIntensityDistributions}

before <- lfqdata$get_Plotter()
before$intensity_distribution_density()

after <- lfqdataNormalized$get_Plotter()
after$intensity_distribution_density()

```


Create a sample of N proteins to speed up computations of models and contrasts.

```{r createDataSubset}
if (SUBSET) {
  N <- 200
  mqdataSubset <- lfqdata$get_sample(size = N, seed = 2020)
  lfqNormSubset <- lfqdataNormalized$get_sample(size = N, seed = 2020)
  lfqNormSubset$hierarchy_counts()
} else {
  mqdataSubset <- lfqdata$clone()
  lfqNormSubset <- lfqdataNormalized$clone()
}
```





```{r R6Ionstar, echo =  FALSE, include = FALSE}
IonstarData <- R6::R6Class(
  "IonstarData",
  public = list(
    data = NULL,
    config = NULL,
    data_N = NULL,
    config_N = NULL,
    
    initialize = function(data, config, data_N, config_N){
      self$data = data
      self$config = config
      self$data_N = data_N
      self$config_N = config_N
    },
    Pep = function(){
      return(list(data = self$data, config = self$config$clone(deep = TRUE)))
    },
    filtered = function(){
      data <- prolfqua:::filter_proteins_by_peptide_count( self$data ,  self$config )$data
      return(list(data = data, config = self$config$clone(deep = TRUE)))
    },
    normalized = function(){
      return(list(data = self$data_N, config = self$config_N))
    },
    subset_normalized = function(){
      return(list(data = self$data_N, config = self$config_N))
    }
  )
)

ionstar <- IonstarData$new(mqdataSubset$data, mqdataSubset$config, 
                           lfqNormSubset$data, lfqNormSubset$config)

#usethis::use_data(data_ionstar, overwrite = TRUE)

```


# Inferring Protein intensities 

We will be using the `LFQDataAggregator` class. To estimate protein intensities using Tukey's median polish we need to use log2 transformed peptide intensities.
The figure below shows the the peptide intensities used for estimation and the protein intensity estimates (black dashed line).

```{r aggregateMedpolish}
lfqAggMedpol <- LFQDataAggregator$new(lfqNormSubset, "protein")
lfqAggMedpol$medpolish()
xx <- lfqAggMedpol$plot()
xx$plots[[1]]
```



```{r storeProteinAggregates, echo = FALSE, include = FALSE}
data_IonstarProtein_subsetNorm <- list(
  data = lfqAggMedpol$lfq_agg$data,
  config = lfqAggMedpol$lfq_agg$config
)

#usethis::use_data(data_IonstarProtein_subsetNorm, overwrite = TRUE)

```

We can also estimate the protein intensities using the top N most abundant peptides. In this case we are using the untransformed protein intensities. The figure below shows the estimated protein intensities.

```{r topNAggregation, echo = TRUE, eval=FALSE}

lfqAggregator <- LFQDataAggregator$new(mqdataSubset, "protein_topN")
lfqAggregator$mean_topN()
topN <- lfqAggregator$plot()
topN$plots[[1]]

```

# Model Fitting

## Model 1 : Fitting a linear model to the protein intensities


```{r buildModelLM}
protLFQ <- lfqAggMedpol$lfq_agg

lmmodel <- "~ dilution."
lmmodel <- paste0(protLFQ$config$table$getWorkIntensity() , lmmodel)

lfqNormSubset$config$table$hierarchyDepth <- 1
modelFunction <- strategy_lm( lmmodel, model_name = "Model")

modLinearProt <- prolfqua::build_model(protLFQ$data, modelFunction)
modLinearProt$anova_histogram()$plot

```

## Model 2: Fitting a mixed effects model to peptide intensities


```{r buildModelLMer}


lmmodel <- "~ dilution. + (1|peptide_Id) + (1|sampleName)"
lmmodel <- paste0(lfqNormSubset$config$table$getWorkIntensity() , lmmodel)
lfqNormSubset$config$table$hierarchyDepth <- 1
modelFunction <- strategy_lmer( lmmodel, model_name = "Model")
modMixedProtLevel <- prolfqua::build_model(lfqNormSubset$data, modelFunction)
modMixedProtLevel$anova_histogram()$plot

```

## Model 3: Fitting peptide level models

```{r buildModelPeptide}

lmmodel <- "~ dilution."

lfqNormSubset$config$table$hierarchyDepth
lfqNormSubset$config$table$hierarchyDepth <- 2

lmmodel <- paste0(lfqNormSubset$config$table$getWorkIntensity() , lmmodel)

modelFunction <- strategy_lm( lmmodel, model_name = "Model")
modLMPepLevel <- prolfqua::build_model(lfqNormSubset$data,
                                       modelFunction,
                                       subject_Id = lfqNormSubset$subject_id())
modLMPepLevel$anova_histogram()$plot

```


#  Computing Contrasts


Once models are fitted contrasts can be computed. The R code below defines all possible contrasts among conditions for the ionstar dataset.

```{r defineContrasts}
DEBUG <- FALSE

Contrasts <- c(
  "dilution_(9/3)_3" =   "dilution.e - dilution.a",
  "dilution_(9/4.5)_2" =   "dilution.e - dilution.b",
  "dilution_(9/6)_1.5" =   "dilution.e - dilution.c",
  "dilution_(9/7.5)_1.2" =   "dilution.e - dilution.d",
  
  "dilution_(7.5/3)_2.5" =   "dilution.d - dilution.a",
  "dilution_(7.5/4.5)_1.6(6)" =   "dilution.d - dilution.b",
  "dilution_(7.5/6)_1.25" =   "dilution.d - dilution.c",
  
  "dilution_(6/3)_2" =   "dilution.c - dilution.a",
  "dilution_(6/4.5)_1.3(3)" =   "dilution.c - dilution.b",
  
  "dilution_(4.5/3)_1.5" =   "dilution.b - dilution.a"
)


tt <- Reduce(rbind, strsplit(names(Contrasts),split = "_"))
tt <- data.frame(tt)[,2:3]
colnames(tt) <- c("ratio" , "expected fold-change")
tt <- tibble::add_column(tt, contrast =  Contrasts, .before = 1)
prolfqua::table_facade(tt, caption = "All possible Contrasts given 5 E. coli dilutions of the Ionstar Dataset")

```


```{r usedContrasts}
relevantContrasts <- c("dilution_(9/7.5)_1.2",
                       "dilution_(7.5/6)_1.25",
                       "dilution_(6/4.5)_1.3(3)",
                       "dilution_(4.5/3)_1.5" )

tt <- Reduce(rbind, strsplit(relevantContrasts,split = "_"))
tt <- data.frame(tt)[,2:3]
colnames(tt) <- c("ratio" , "expected fold-change")
tt <- tibble::add_column(tt, contrast =  Contrasts[names(Contrasts) %in% relevantContrasts], .before = 1)
prolfqua::table_facade(tt, caption = "Contrasts used for benchmark.")
relevantContrasts <- Contrasts[names(Contrasts) %in% relevantContrasts]

```


There are, as of today, four contrasts classes in the package _prolfqua_
- 'ContrastSimpleImputed' : contrast computation with imputation of fold changes and t-statistic estimation using pooled variances.
- 'Contrast'  : uses Wald test, 
- 'ContrastModerated' : applies variance moderation:, 
- 'ContrastROPECA' implements estimate and p-value aggregation


## Contrasts with Imputation

In order to estimate fold-changes, statistics and p-values of proteins for which linear models could not be fitted because of an excess of missing measurements, the following procedure is applied. The mean intensity of a protein over all samples in a condition is computed. For the proteins with no observation in one condition, we imputed their intensities using the mean of the 10% smallest average peptide intensities. The standard deviation is estimated in the condition where the peptide is observed.


```{r contrastImputed}
contrImp <- prolfqua::ContrastsSimpleImpute$new(protLFQ, relevantContrasts)
pl <- contrImp$get_Plotter()
bb <- contrImp$get_contrasts()
plc <- contrImp$get_Plotter()
plc$volcano()
plc$histogram()
```

```{r include = FALSE}
dummy <- protLFQ$clone()
dummy$data <- protLFQ$data %>% filter(protein_Id == "sp|A0A0U1RRL7|MMPOS_HUMAN~A0A0U1RRL7")
#prolfqua::ContrastsSimpleImpute$debug("get_contrasts")
contrImp2 <- prolfqua::ContrastsSimpleImpute$new(dummy, relevantContrasts)

#debug(summarize_stats)
#var = summarize_stats(dummy$data, dummy$config, all = FALSE)
x2 <- contrImp2$get_contrasts()

```


```{r benchmarkImputed}
allContrasts <- list()
allContrasts$imputation <- contrImp$get_contrasts()
ttd <- prolfqua::ionstar_bench_preprocess(contrImp$get_contrasts())

benchmark_imputation <- make_benchmark(ttd$data,
                                   model_description = "med. polish and lm. imputation",
                                   model_name = "prot_imputation",
                                   FDRvsFDP = list(list(sc = "FDR", desc = FALSE))
)

benchmark_imputation$plot_ROC()
benchmark_imputation$plot_FDRvsFDP()

prolfqua::table_facade(benchmark_imputation$smc$summary, caption = "Nr of proteins with Nr of estimated contrasts.")


allBenchmarks <- list()

allBenchmarks$benchmark_imputation <- benchmark_imputation

```


## Contrasts from Model 1

```{r contrastModel1}
dim(modLinearProt$modelDF)

contrProt <- prolfqua::Contrasts$new(modLinearProt, relevantContrasts)
pl <- contrProt$get_Plotter()
pl$volcano()
pl$histogram()
```


```{r benchmarkModel1}
allContrasts$Prot <- contrProt$get_contrasts()

ttd <- prolfqua::ionstar_bench_preprocess(contrProt$get_contrasts())
ttd$data$protein_Id %>% unique %>% length

benchmark_Prot <- make_benchmark(ttd$data,
                                   model_description = "med. polish and lm",
                                   model_name = "prot_med_lm"
)
prolfqua::table_facade(benchmark_Prot$smc$summary, caption = "Nr of proteins with Nr of estimated contrasts.")


#benchmark_Prot$plot_score_distribution()
benchmark_Prot$plot_FDRvsFDP()
allBenchmarks$benchmark_Prot <- benchmark_Prot
```


### Adding Moderation

```{r contrastModel1Moderation}

contrProtModerated <- ContrastsModerated$new(contrProt)
contrProtModerated$get_Plotter()$volcano()
contrProtModerated$get_contrasts()$protein_Id %>% unique %>% length

```


```{r benchmarkModel1Moderation}
allContrasts$ProtModerated <- contrProtModerated$get_contrasts()

ttd <- prolfqua::ionstar_bench_preprocess(contrProtModerated$get_contrasts())

benchmark_ProtModerated <- make_benchmark(ttd$data,
                                   model_description = "med. polish and lm moderated",
                                   model_name = "prot_med_lm_moderated")

#benchmark_ProtModerated$plot_score_distribution()
prolfqua::table_facade(benchmark_ProtModerated$smc$summary,
                       caption = "Nr of proteins with Nr of estimated contrasts.")

benchmark_ProtModerated$plot_FDRvsFDP()
allBenchmarks$benchmark_ProtModerated <- benchmark_ProtModerated

```

## Contrasts from Model 2

```{r contrastModel2}
contrProtMixed <- prolfqua::Contrasts$new(modMixedProtLevel, relevantContrasts, modelName = "WaldTestMixed")

pl <- contrProtMixed$get_Plotter()
pl$volcano()
pl$histogram()

pl <- contrProtMixed$get_contrasts()
pl$protein_Id %>% unique %>% length
```


```{r benchmarkModel2}
allContrasts$contrProtMixed <- contrProtMixed$get_contrasts()
ttd <- prolfqua::ionstar_bench_preprocess(contrProtMixed$get_contrasts())
benchmark_mixed <- make_benchmark(ttd$data,
                                   model_description = "mixed effect model",
                                   model_name = "prot_mixed_effect"
)
benchmark_mixed$complete(FALSE)
prolfqua::table_facade(benchmark_mixed$smc$summary, caption = "Nr of proteins with Nr of estimated contrasts.")

#benchmark_mixed$plot_score_distribution()
benchmark_mixed$plot_FDRvsFDP()
allBenchmarks$benchmark_mixed <- benchmark_mixed
```

### Adding Moderation

Since moderation requires a good degrees of freedom estimate to estimate the prior degrees of freedom we examine the denominator degrees of freedom produced by the methods implemented in lmerTest (see Histogram).

```{r distributionOfDFModel2, fig.cap="Histogram of degrees of freedom for mixed model"}
ctr <- contrProtMixed$get_contrasts()
df <- ctr$df
df[df > 59] <- 60
range(df)
hist(df, breaks = 100, xlim = c(0,61))
```


```{r contrastModel2Moderation}

contrProtMixedModerated <- ContrastsModerated$new(contrProtMixed)
contrProtMixedModerated$get_Plotter()$volcano()
```


```{r benchmarkModel2Moderation}
allContrasts$contrProtMixedModerated <- contrProtMixedModerated$get_contrasts()

ttd <- prolfqua::ionstar_bench_preprocess(contrProtMixedModerated$get_contrasts())

benchmark_mixedModerated <- make_benchmark(ttd$data,
                                   model_description = "mixed effect model moderated",
                                   model_name = "prot_mixed_effect_moderated")

prolfqua::table_facade(benchmark_mixedModerated$smc$summary, caption = "Nr of proteins with Nr of estimated contrasts.")

#benchmark_mixedModerated$plot_score_distribution()
benchmark_mixedModerated$plot_FDRvsFDP()
allBenchmarks$benchmark_mixedModerated <- benchmark_mixedModerated
```


## Protein level contrasts from Peptide models

To estimate regulation probabilities using the ROPECA approach we can chain the contrast computation methods. First we compute contrasts on peptide level, than we moderated the variance, t-statistics and p-values and finally we aggregate the fold change estimates and p-values.

```{r contrastROPECA}
contrROPECA <- prolfqua::Contrasts$new(modLMPepLevel, relevantContrasts) %>%
  ContrastsModerated$new() %>%
  ContrastsROPECA$new()

contrROPECA$get_Plotter()$volcano()
contrROPECA$get_Plotter()$histogram()

```


```{r benchmarkROPECA}
#View(contrROPECA$get_contrasts())
allContrasts$contrProtMixedModerated <- contrProtMixedModerated$get_contrasts()

ttd <- prolfqua::ionstar_bench_preprocess(contrROPECA$get_contrasts())
benchmark_ropeca <- make_benchmark(ttd$data,
                                toscale = c("beta.based.significance"),
                                benchmark = list(
                                  list(sc = "estimate", desc = TRUE),
                                  list(sc = "statistic", desc = TRUE),
                                  list(sc = "scaled.beta.based.significance", desc = TRUE)
                                ),  
                                model_description = "Ropeca",
                                model_name = "ropeca",
                                FDRvsFDP = list(list(sc = "FDR.beta.based.significance", desc = FALSE))
)


prolfqua::table_facade(benchmark_ropeca$smc$summary, caption = "Nr of proteins with Nr of estimated contrasts.")

benchmark_ropeca$plot_ROC()
benchmark_ropeca$plot_FDRvsFDP()
allBenchmarks$benchmark_ropeca <- benchmark_ropeca

```


## Merging contrasts


```{r}

merged <- prolfqua::addContrastResults(prefer = contrProtMixedModerated, add = contrImp)
allContrasts$merged <- merged
saveRDS(allContrasts, file = "../inst/Benchresults/allContrasts.RDS")

ttd <- prolfqua::ionstar_bench_preprocess(merged)
benchmark_merged <- make_benchmark(ttd$data,
                                   model_description = "merge of prot moderated and imputed",
                                   model_name = "prot_merged")

prolfqua::table_facade(benchmark_merged$smc$summary, caption = "Nr of proteins with Nr of estimated contrasts.")

#benchmark_mixedModerated$plot_score_distribution()
benchmark_merged$plot_FDRvsFDP()
benchmark_merged$plot_ROC()
allBenchmarks$benchmark_merged <- benchmark_merged
saveRDS(allBenchmarks, file = "../inst/Benchresults/allBenchmarkds.RDS")
```

# Benchmarking

The table below summarizes the contrast estimates produced which will be benchmarked.

|                    | Model                   | Contrast   | Moderation   | Aggregation  |
|--------------------|:-----------------------:|------------|--------------|--------------|
| Protein Intensity  |  lm                     | o          |  o           |              |
| Protein Intensity Imputed | pooled variance  | o          |  o           |              |
| Peptide Intensity  |  lmer                   | o          |  o           |              |
| Peptide Intensity  |  lm                     |            |              |  o           |


```{r fig.cap="Partial area under the ROC curve at 10% FPR."}
ttt <- sapply(allBenchmarks, function(x){x$complete(FALSE)})
res <- map_df(allBenchmarks, function(x){x$pAUC()})
resAllB <- res %>% filter(contrast == "all")

bb <- resAllB %>% mutate(whatfix = case_when(what == "scaled.beta.based.significance" ~ "scaled.p.value", TRUE ~ what))

ggplot(bb, aes(x = Name, y = pAUC_10)) +
  geom_bar(stat = "identity") +
  facet_wrap(~whatfix)  + 
  coord_cartesian(ylim = c(65,82.5)) + 
  theme_minimal() + 
  theme(axis.text.x = element_text(angle = -90, vjust = 0.5))

```

```{r fig.caption="Partial area under the ROC curve at 20% FPR."}
ggplot(bb, aes(x = Name, y = pAUC_20)) +
  geom_bar(stat = "identity") +
  facet_wrap(~whatfix)  +
  coord_cartesian(ylim = c(80,87.5)) + 
  theme_minimal() + 
  theme(axis.text.x = element_text(angle = -90, vjust = 0.5))  

```


### Look at the nr of estimated contrasts.

```{r fig.cap="NR of estimated contrasts", fig.width=5, fig.height=5}
dd <- map_df(allBenchmarks, function(x){res <- x$smc$summary; res$name <- x$model_name;res})
dd$nr_missing <- as.integer(dd$nr_missing) - 1
dd <- dd %>% mutate(nrcontrasts = protein_Id * (4-nr_missing))
dds <- dd %>% group_by(name) %>% summarize(nrcontrasts = sum(nrcontrasts))

dds %>% ggplot(aes(x = name, y = (nrcontrasts - min(nrcontrasts)))) + 
  geom_bar(stat="identity") + 
  theme(axis.text.x = element_text(angle = -90, vjust = 0.5)) +
  geom_text(aes(label=nrcontrasts), position=position_dodge(width=0.9), vjust=-0.25)

```

### Plot FDR vs FDP

```{r FDRfdp, fig.cap = "Compare FDR estimate with false discovery proportion (FDP).", fig.width=8, fig.height=8}
dd <- map_df(allBenchmarks, function(x){res <- x$get_FDRvsFDP(); res$name <- x$model_name;res})

dd %>% ggplot(aes(y = FDP_,  x  = scorecol )) + 
  geom_line(aes(color = model_name)) +
  facet_wrap(~contrast) + 
   geom_abline(intercept = 0, slope = 1, color=2)

  
  


```


