---
title: "Comparing Two Groups"
author: "Witold Wolski"
date: "`r format(Sys.time(), '%d %B, %Y')`"
output: html_document
vignette: >
  %\VignetteIndexEntry{Comparing Two Groups} 
  %\VignetteEncoding{UTF-8}
  %\VignetteEngine{knitr::rmarkdown}
editor_options: 
  chunk_output_type: console
---


```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE, message = FALSE, warning = FALSE)
conflicted::conflict_prefer("filter", "dplyr")
```


# Purpose
In this vignette we demonstrate how two conditions or groups can be compared and statistically evaluated. We will again use the Ionstar dataset processed with MaxQuant as LFQ experiment with MBR option. After some standard preprocessing we compare in the end the replicates with the different dilutions with each other. The output of such a two group comparison is usually a difference in the mean intensities for quantified proteins (log2 fold-change) along with some statistical parameters such as a p-value or a FDR. 

# Loading protein abundances from MaxQuant proteinGroups.txt

```{r createConfig}
library(prolfqua)
```

Specify the path to the MaxQuant `proteinGroups.txt` file. The function `tidyMQ_ProteinGroups` will read the `proteinGroups.txt` file and convert it into a tidy table

```{r LoadDataAndConfigure}
datadir <- file.path(find.package("prolfqua") , "samples/maxquant_txt")
inputMQfile <-  file.path(datadir, "tiny2.zip")
inputAnnotation <- file.path(datadir, "annotation_Ionstar2018_PXD003881.xlsx")
startdata <- prolfqua::tidyMQ_ProteinGroups(inputMQfile)
```


Read the sample annotation. The sample annotation must contain the `raw.file` name and the explanatory variables of your experiment, e.g. treatment, timepoint, genetic background, or other `factors` which you would like to check for confounding effects.


```{r readAnnotation}
annotation <- readxl::read_xlsx(inputAnnotation)
head(annotation)
```

Merge the annotation with quantitative data using `inner_join` joining by `raw.file`.

```{r addAnnotationToData}
startdata <- dplyr::inner_join(annotation, startdata, by = "raw.file")

```

We remove all proteins identified only by a single peptide.

```{r filterForAtLeastTwoPeptides}
startdata <- dplyr::filter(startdata, nr.peptides > 1)
```

Then you need to _tell_ `prolfqua` which columns in the data frame contain what information. You do it using the `AnalysisTableAnnotation` class.

```{r setupConfigs}
atable <- AnalysisTableAnnotation$new()
```

The `AnalysisTableAnnotation` has the following fields that always need to be populated:

- fileName
- hierarchy
- factors
- workingIntensity


The `fileName` is the column with the raw file names, however for labelled TMT experiments, it can be used to hold the name of the TMT channel.

```{r specifyRawFile}
atable$fileName = "raw.file"
```

The `hierarchy` field describes the structure of the MS data e.g, 

- protein 
- peptides
- modified peptides
- precursor

In case of the MQ proteinGroups file we have the data already on protein level.

```{r specifyProteinID}
atable$hierarchy[["protein_Id"]] <- c("proteinID")

```

In addition you need to describe the `factors` of the analysis, i.e, the column containing the explanatory variables. 
By setting the `factorDepth` variable to $1$ we include here only the "dilution." variable into the modelling.

```{r specifyFactors}
atable$factors[["dilution."]] = "sample"
```

We also need to specify the column containing the protein abundances.

```{r specifyIntensity}
atable$set_response("mq.protein.intensity")
```

Finally we create the `AnalysisConfiguration` which needs the `AnalysisTableAnnotation` we just created and the `AnalysisParameters`. 

```{r createAnalysisConfig}
config <- AnalysisConfiguration$new(atable)
adata <- setup_analysis(startdata, config)
```

Create the `LFQData` class instance and remove zeros from data (MaxQuant encodes missing values with zero).

```{r removeSmallIntensities}
lfqdata <- prolfqua::LFQData$new(adata, config)
lfqdata$remove_small_intensities()
lfqdata$factors()
```

You can always convert the data into wide format.

```{r showWide, eval = TRUE}
lfqdata$to_wide()$data[1:3,1:7]
```

## Visualization of not normalized data

After this first setting up of the analysis we show now how to normalize the proteins and the effect of normalization. Furthermore we use some functions to visualize the missing values in our data. 

```{r getPlotter}
lfqplotter <- lfqdata$get_Plotter()
density_nn <- lfqplotter$intensity_distribution_density()
```

### Visualization of missing data

```{r makeMissingHeatmap, fig.cap="Heatmap where missing proteins (zero in case of MaxQuant reported intensities), black - missing protein intensities, white - present"}
lfqplotter$NA_heatmap()
```

```{r missignessPerGroup, fig.cap="# of proteins with 0,1,...N missing values"}
lfqdata$get_Summariser()$plot_missingness_per_group()
```

```{r missignessHistogram, fig.cap="Intensity distribution of proteins depending on # of missing values"}
lfqplotter$missigness_histogram()
```

### Computing standard deviations, mean and CV.

Other important statistics such as coefficient of variation, means and standard deviations can be easily calculated using the `get_Stats` function and visualized with a violin plot.

```{r PlotCVDistributions, fig.cap="Violin plots of CVs in the different groups and among all groups"}
stats <- lfqdata$get_Stats()
stats$violin()
prolfqua::table_facade( stats$stats_quantiles()$wide, paste0("quantile of ",stats$stat ))
```


```{r plotCVsplitbyMedianIntensity, fig.cap="Distribution of CV's for top 50% and bottom 50% proteins by intensity."}
stats$density_median()

```


## Normalize protein intensities and show diagnostic plots

We normalize the data by $\log_2$ transforming and then $z-scaling$.

```{r normalizedata}
lt <- lfqdata$get_Transformer()
transformed <- lt$log2()$robscale()$lfq
transformed$config$table$is_response_transformed

```


```{r genplotNorm, fig.cap="Normalized intensities."}
pl <- transformed$get_Plotter()
density_norm <- pl$intensity_distribution_density()
```



```{r showIntensityDistributions, fig.cap="Distribution of intensities before and after normalization."}
gridExtra::grid.arrange(density_nn, density_norm)
```

```{r plotScatterMatrix, fig.cap = "Scatterplot matrix"}
pl$pairs_smooth()
```

```{r createHeatmap}
p <- pl$heatmap_cor()
```

```{r plotHeatmap, fig.cap="Heatmap, Rows - proteins, Columns - samples", fig.align=5, fig.height=5}
p
```

# Fitting a linear model 

For fitting linear models to the transformed intensities for all our proteins we have to first specify the model function and define the contrasts that we want to calculate.

```{r defineModelAndContrasts}
transformed$config$table$get_response()
formula_Condition <-  strategy_lm("transformedIntensity ~ dilution.")

# specify model definition
modelName  <- "Model"
unique(transformed$data$dilution.)
Contrasts <- c("dila - dilb" = "dilution.a - dilution.b",
"AvsC" = "dilution.a - dilution.c")

```

Here we have to build the model for each protein.

```{r buildModel}
mod <- prolfqua::build_model(
  transformed$data,
  formula_Condition,
  subject_Id = transformed$config$table$hierarchy_keys() )

```

In this plot we can see what factors in our model are mostly responsible for the adjusted p-values calculated from an analysis of variance.

```{r showANOVA, fig.cap="Distribtuion of adjusted p-values (FDR)"}
mod$anova_histogram("FDR.Pr..F.")
```

One also look what proteins do show different abundances in any of our five dilutions by looking at the FDR values of the analysis of variane.

```{r filterDataForFDR}
aovtable <- mod$get_anova()
head(aovtable)
dim(aovtable)
xx <- aovtable |> dplyr::filter(FDR.Pr..F. < 0.2)
signif <- transformed$get_copy()
signif$data <- signif$data |> dplyr::filter(protein_Id %in% xx$protein_Id)
hmSig <- signif$get_Plotter()$heatmap()

```

```{r showSigHeatmap, fig.cap="Heatmap for proteins with FDR < 0.2 in the analysis of variance"}
hmSig

```

# Compute contrasts

Next we do calculate the statistics for our defined contrasts for all the proteins.
For this we can use the `Contrasts` function.

```{r computeContrasts}
contr <- prolfqua::Contrasts$new(mod, Contrasts)
v1 <- contr$get_Plotter()$volcano()
```

Alternatively, we can moderate the variance and using the Experimental Bayes method implemented in `ContrastsModerated`.

```{r computeContrastsModerated}
contr <- prolfqua::ContrastsModerated$new(contr)
contrdf <- contr$get_contrasts()
```

In the next figure it can be seen WEWinputNEEDED.

```{r plotVolcanos, fig.cap="Volcano plot, Left panel - no moderation, Right panel - with moderation."}
plotter <- contr$get_Plotter()
v2 <- plotter$volcano()
gridExtra::grid.arrange(v1$FDR,v2$FDR, ncol = 1)
```

```{r showMAplot, fig.cap="MA plot showing the dependency of mean abuncance with respect to the difference"}
plotter$ma_plotly()
```


```{r checkProteinsInOutput}
#myProteinIDS <- c("sp|Q12246|LCB4_YEAST",  "sp|P38929|ATC2_YEAST",  "sp|Q99207|NOP14_YEAST")
myProteinIDS <- c("sp|P0AC33|FUMA_ECOLI",  "sp|P28635|METQ_ECOLI",  "sp|Q14C86|GAPD1_HUMAN")
dplyr::filter(contrdf, protein_Id %in% myProteinIDS)
```


# Contrasts with missing value imputation

Use this method if there proteins with no observations in one of the groups. With the `ContrastsSimpleImpute` function, we can estimate difference in mean for proteins that are not observed in one group or condition. For this we are using the average expression at percentile 0.05 of the group where the protein is not quantified. 

```{r computeMissing, eval=TRUE}
mC <- ContrastsSimpleImpute$new(lfqdata = transformed, contrasts = Contrasts)
colnames(mC$get_contrasts())

```

Finally we are merging the results and give priority to the results where we do not have missing values in one group.

```{r mergeResults, fig.cap="Volcano plots for the two contrasts with missing value imputation from the group_average model."}

merged <- prolfqua::merge_contrasts_results(prefer = contr,add = mC)$merged
plotter <- merged$get_Plotter()
tmp <- plotter$volcano()
tmp$FDR

```

Look at proteins which could not be fitted using the linear model, if any.

```{r mergedMore}
merged <- prolfqua::merge_contrasts_results(prefer = contr,add = mC)

moreProt <- transformed$get_copy()
moreProt$data <- moreProt$data |> dplyr::filter(protein_Id %in% merged$more$contrast_result$protein_Id)
moreProt$get_Plotter()$raster()
```

# GSEA Analyis

We can rank the proteins based on the log2FC or the t-statistic and subject them them to gene set enrichment analysis (see GSEA).

This example will run only if the following packages are installed on you machine:

- clusterProfiler (Bioconductor)
- org.Sc.sgd.db (Bioconductor)
- prora (github.com/protviz/prora)


```{r prepForGSEA}
#evalAll <- require("clusterProfiler") & require("org.Sc.sgd.db") & require("prora")
evalAll <- require("clusterProfiler") & require("org.Sc.sgd.db2") & require("prora")

```

```{r clusterProfiler, eval=evalAll}
library(clusterProfiler)
library(org.Sc.sgd.db)

bb <- prolfqua::get_UniprotID_from_fasta_header(merged$merged$get_contrasts(),
                                             idcolumn = "protein_Id")
bb <- prora::map_ids_uniprot(bb)
ranklist <- bb$statistic
names(ranklist) <- bb$P_ENTREZGENEID
res <- clusterProfiler::gseGO(
  sort(ranklist, decreasing = TRUE),
  OrgDb = org.Sc.sgd.db,
  ont = "ALL")
```

```{r ridgeplot, fig.cap="ridgeplot", eval = evalAll}
ridgeplot( res )
```

```{r dotplot, fig.cap = "Dotplot", eval = evalAll}
dotplot(res , showCategory = 30)
```

```{r upsetplot, fig.cap="Upset Plot", eval = evalAll}
enrichplot::upsetplot(res)
```
