---
title: "Comparing Two Groups"
author: "Witold Wolski"
date: "14/12/2020"
output: pdf_document
vignette: >
  %\VignetteIndexEntry{Comparing Two Groups} 
  %\VignetteEncoding{UTF-8}
  %\VignetteEngine{knitr::rmarkdown}
editor_options: 
  chunk_output_type: console
---


```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE, message = FALSE, warning = FALSE)
```


# Loading data

Create configuration for MQ peptide file or use function `create_config_MQ_peptide`

```{r createConfig}
library(prolfqua)
library(tidyverse)

```


```{r LoadDataAndConfigure}
datadir <- file.path(find.package("prolfquaData") , "quantdata")
inputMQfile <-  file.path(datadir, "MAXQuant_ComboCourse_p2691_March_2018_WU183012.zip")
inputAnnotation <- file.path(datadir, "annotation_ComboCourse_p2691_March_2018_WU183012.xlsx")

startdata <- prolfqua::tidyMQ_ProteinGroups(inputMQfile)
ncol(startdata)
str(startdata)
annotation <- readxl::read_xlsx(inputAnnotation)
annotation$experiment = "p2691"
#View(annotation)
ncol(annotation)


startdata <- inner_join(annotation, startdata, by = "raw.file")
ncol(startdata)
str(startdata)
```

For the sample size estimation we will use bio-reps from one condition. Also remove all proteins identified only by a single peptide.


```{r}
startdata <- filter(startdata, nr.peptides > 1)

```

Telling `prolfqua` at which columns to look.

The `hierarchy` describes the structure of the MS data e.g, 
  - protein 
  - peptides
  - modified peptides
  - precursor

In this case we have only protein level measurements.

In addition you need to describe the factors of the analysis, i.e, the column containing the explanatory variables.
Then you need to add the intensity column.

```{r setupConfigs}

atable <- AnalysisTableAnnotation$new()
atable$fileName = "raw.file"
atable$hierarchy[["protein_Id"]] <- c("majority.protein.ids")

atable$hierarchyDepth <- 1
atable$setWorkIntensity("mq.protein.intensity")

anaparam <- AnalysisParameters$new()
config <- AnalysisConfiguration$new(atable, anaparam)


config$table$factors[["Condition_"]] = "condition"
config$table$factors[["run"]] = "Run_ID"
config$table$factorDepth <- 1

adata <- setup_analysis(startdata, config)
str(adata)
```


Create the LFQData class instance, remove zeros from data.

```{r filterdata}
lfqdata <- LFQData$new(adata, config)
lfqdata$remove_small_intensities()
lfqdata$factors()
```

You can convert the data into wide format.

```{r}
lfqdata$to_wide()$data
```


## Visualization of not normalized data

```{r makeMissingHeatmap}
lfqplotter <- lfqdata$get_Plotter()
lfqplotter$intensity_distribution_density()
```


### Visualization of missing data


```{r fig.cap="# of proteins with 0,1,...N missing values"}
lfqplotter$missingness_per_condition()
```

```{r fig.cap="Intensity distribution of proteins depending on # of missing values"}
lfqplotter$missigness_histogram()
lfqplotter$intensity_distribution_density()
```

# Computing standard deviations, mean and CV.


```{r PlotCVDistributions}
stats <- lfqdata$get_Stats()
stats$violin()
prolfqua::table_facade( stats$stats_quantiles()$wide, paste0("quantile of ",stats$stat ))
```


```{r fig.cap="Distribution of CV's"}
stats$density_median()

```


## Normalize protein intensities

We normalize the data by $\log2$ transforming and then z-scaling.

```{r normalizedata}
lt <- lfqdata$get_Transformer()
transformed <- lt$log2_robscale()
transformed$config$table$is_intensity_transformed

```


```{r plotDensities, fig.cap="Normalized intensities."}
pl <- transformed$get_Plotter()
pl$intensity_distribution_density()
```

```{r plotScatterMatrix, fig.cap = "Scatterplot matrix"}
pl$pairs_smooth()
```

```{r createHeatmap}
p <- pl$heatmap_cor()
```

```{r plotheatmap, fig.cap="Heatmap, Rows - proteins, Columns - samples", fig.align=5, fig.height=5}
p
```

# Model specification

```{r}
transformed$config$table$getWorkIntensity()

formula_Condition <-  make_custom_model_lm("transformedIntensity ~ Condition_")

# specify model definition
modelName  <- "Model"
unique(transformed$data$Condition_)
Contrasts <- c("Glucose - Ethanol" = "Condition_Glucose - Condition_Ethanol")
Contrasts <- c("GvsE" = "Condition_Glucose - Condition_Ethanol")


```



```{r buildModel}
mod <- prolfqua::build_model(
  transformed$data,
  formula_Condition,
  subject_Id = transformed$config$table$hierarchyKeys() )

```



# Compute contrasts

```{r}
contr <- ContrastsModerated$new(mod, Contrasts)
#contr$get_contrasts_sides()
contrdf <- contr$get_contrasts()
dim(contrdf)
#View(contrdf)
```



```{r}
plotter <- contr$get_Plotter()
plotter$volcano()
plotter$ma_plot()
plotter$ma_plotly()
```


```{r}
myProteinIDS <- c("sp|Q12246|LCB4_YEAST",  "sp|P38929|ATC2_YEAST",  "sp|Q99207|NOP14_YEAST")
filter(contrdf, protein_Id %in% myProteinIDS)

```


# Missing values

```{r}
mC <- ContrastsSimpleImpute$new(lfqdata = transformed, contrasts = Contrasts)
dim(mC$get_contrasts())

```

