---
title: "Comparing Two Groups"
author: "Witold Wolski"
date: "`r format(Sys.time(), '%d %B, %Y')`"
output: html_document
vignette: >
  %\VignetteIndexEntry{Comparing Two Groups} 
  %\VignetteEncoding{UTF-8}
  %\VignetteEngine{knitr::rmarkdown}
editor_options: 
  chunk_output_type: console
---


```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE, message = FALSE, warning = FALSE)
conflicted::conflict_prefer("filter", "dplyr")
```


# Loading data

Load the required packages.

```{r createConfig}
library(prolfqua)
library(tidyverse)

```

Specify the path to the MaxQuant proteinGroups.txt file. The function `tidyMQ_ProteinGroups` will read the `proteinGroups.txt` file and convert it into a long format.

```{r LoadDataAndConfigure}
datadir <- file.path(find.package("prolfquaData") , "quantdata")
inputMQfile <-  file.path(datadir, "MAXQuant_ComboCourse_p2691_March_2018_WU183012.zip")
inputAnnotation <- file.path(datadir, "annotation_ComboCourse_p2691_March_2018_WU183012.xlsx")

startdata <- prolfqua::tidyMQ_ProteinGroups(inputMQfile)
```


Read the sample annotation. The sample annotation must contain the `raw.file` name and the explanatory variables of your experiment, e.g. treatment, timepoint, genetic background, or other factors which you would like to check for confounding.


```{r}
annotation <- readxl::read_xlsx(inputAnnotation)
head(annotation)
```

Merge the annotation with quantitative data using `inner_join` joining by `raw.file`.

```{r}
startdata <- dplyr::inner_join(annotation, startdata, by = "raw.file")
```

We remove all proteins identified only by a single peptide.


```{r}
startdata <- filter(startdata, nr.peptides > 1)
```


Then you need to _tell_ `prolfqua` which columns in the data frame contain what information. You do it using the `AnalysisTableAnnotation` class.

```{r setupConfigs}
atable <- AnalysisTableAnnotation$new()
```

The AnalysisTableAnnotation has the following fields that need to be populated:
- fileName
- hierarchy
- factors
- workingIntensity,
and we will discuss in more detail next.


The `fileName` is the column with the raw file names. 
```{r}
atable$fileName = "raw.file"
```


The `hierarchy` field describes the structure of the MS data e.g, 
- protein 
- peptides
- modified peptides
- precursor
In case of the MQ proteinGroups file we have data on protein level.

```{r}
atable$hierarchy[["protein_Id"]] <- c("proteinID")

```


In addition you need to describe the `factors` of the analysis, i.e, the column containing the explanatory variables. We include here the "Run_ID". It will be visualized int some of the plots. By setting the factorDepth variable to $1$ we include only the "Condition_" variable into the modelling.


```{r}
atable$factors[["Condition_"]] = "condition"
atable$factors[["run"]] = "Run_ID"
atable$factorDepth <- 1
```


```{r}
startdata$Run_ID <- as.integer(startdata$Run_ID)
```


We also need to specify the intensity column.

```{r}
atable$setWorkIntensity("mq.protein.intensity")
```

Finally we create the `AnalysisConfiguration` which needs the `AnalysisTableAnnotation` we just created and the `AnalysisParameters`. 
```{r}
config <- AnalysisConfiguration$new(atable)

adata <- setup_analysis(startdata, config)

```


Create the LFQData class instance and remove zeros from data (MaxQuant encodes missing values with zero).

```{r filterdata}
lfqdata <- LFQData$new(adata, config)
lfqdata$remove_small_intensities()
lfqdata$factors()
```

You can convert the data into wide format.

```{r eval = FALSE}
lfqdata$to_wide()$data
wr <- lfqdata$get_Writer()
wr$write_wide(".")
```


## Visualization of not normalized data



```{r makeMissingHeatmap}
lfqplotter <- lfqdata$get_Plotter()
density_nn <- lfqplotter$intensity_distribution_density()
```


### Visualization of missing data


```{r fig.cap="# of proteins with 0,1,...N missing values"}
lfqplotter$missingness_per_condition()
```

```{r fig.cap="Intensity distribution of proteins depending on # of missing values"}
lfqplotter$missigness_histogram()
```

### Computing standard deviations, mean and CV.


```{r PlotCVDistributions}
stats <- lfqdata$get_Stats()
stats$violin()
prolfqua::table_facade( stats$stats_quantiles()$wide, paste0("quantile of ",stats$stat ))
```


```{r fig.cap="Distribution of CV's"}
stats$density_median()

```


## Normalize protein intensities

We normalize the data by $\log_2$ transforming and then $z-scaling$.

```{r normalizedata}
lt <- lfqdata$get_Transformer()
transformed <- lt$log2()$robscale()$lfq
transformed$config$table$is_intensity_transformed

```


```{r plotDensities, fig.cap="Normalized intensities."}
pl <- transformed$get_Plotter()
density_norm <- pl$intensity_distribution_density()
```



```{r fig.cap="Distribution of intensities before and aftern normalization."}
gridExtra::grid.arrange(density_nn, density_norm)
```

```{r plotScatterMatrix, fig.cap = "Scatterplot matrix"}
pl$pairs_smooth()
```

```{r createHeatmap}
p <- pl$heatmap_cor()
```

```{r plotheatmap, fig.cap="Heatmap, Rows - proteins, Columns - samples", fig.align=5, fig.height=5}
p
```

# Fitting a linear model 


```{r}
transformed$config$table$getWorkIntensity()

formula_Condition <-  strategy_lm("transformedIntensity ~ Condition_")

# specify model definition
modelName  <- "Model"
unique(transformed$data$Condition_)
Contrasts <- c("Glucose - Ethanol" = "Condition_Glucose - Condition_Ethanol")
Contrasts <- c("GvsE" = "Condition_Glucose - Condition_Ethanol")


```



```{r buildModel}
mod <- prolfqua::build_model(
  transformed$data,
  formula_Condition,
  subject_Id = transformed$config$table$hierarchyKeys() )

```


```{r fig.cap="Distribtuion of "}
mod$anova_histogram("FDR.Pr..F.")
```

```{r}
aovtable <- mod$get_anova()
head(aovtable)
dim(aovtable)
xx <- aovtable %>% filter(FDR.Pr..F. < 0.2)
dim(xx)
signif <- transformed$get_copy()
signif$data <- signif$data %>% filter(protein_Id %in% xx$protein_Id)
hmSig <- signif$get_Plotter()$heatmap()

```

```{r}
hmSig

```

# Compute contrasts

```{r contrModerated}
contr <- prolfqua::Contrasts$new(mod, Contrasts)
v1 <- contr$get_Plotter()$volcano()
```

Moderate the variance and using the Experimental Bayes method implmented in `ContrastModerated`.

```{r}
contr <- prolfqua::ContrastsModerated$new(contr)
#contr$get_contrasts_sides()
contrdf <- contr$get_contrasts()
#View(contrdf)
```



```{r plotVolcanos, fig.cap="Volcano plot, Left panel - no moderation, Right panel - with moderation."}
plotter <- contr$get_Plotter()
v2 <- plotter$volcano()
gridExtra::grid.arrange(v1$FDR,v2$FDR, ncol = 1)
```

```{r fig.cap="MA plot"}
plotter$ma_plotly()
```


```{r checkProteinsInOutput}
myProteinIDS <- c("sp|Q12246|LCB4_YEAST",  "sp|P38929|ATC2_YEAST",  "sp|Q99207|NOP14_YEAST")
filter(contrdf, protein_Id %in% myProteinIDS)

```


# Contrasts with missing value imputation

```{r computeMissing, eval=TRUE}
mC <- ContrastsSimpleImpute$new(lfqdata = transformed, contrasts = Contrasts)
colnames(mC$get_contrasts())

```

Merging both results.

```{r}

merged <- prolfqua::addContrastResults(prefer = contr,add = mC)$merged

merged$modelName <- as.factor(merged$modelName)
plotter <- Contrasts_Plotter$new(contrastDF = merged , subject_Id = contr$subject_Id)

tmp <- plotter$volcano()
tmp$FDR


```

Look at proteins which could not be fitted using the linear model.

```{r}
merged <- prolfqua::addContrastResults(prefer = contr,add = mC)
names(merged)

moreProt <- transformed$get_copy()
moreProt$data <- moreProt$data %>% filter(protein_Id %in% merged$more$protein_Id)
moreProt$get_Plotter()$raster()
```

# GSEA Analyis

We can rank the proteins based on the log2FC or the t-statistic and subject them them to gene set enrichment analysis (see GSEA).

This example will run only if the following packages are installed on you machine:

- clusterProfiler (Bioconductor)
- org.Sc.sgd.db (Biocondcutor)
- prora (github.com/protviz/prora)


```{r}
evalAll <- require("clusterProfiler") & require("org.Sc.sgd.db") & require("prora")

```

```{r eval=evalAll}
library(clusterProfiler)
library(org.Sc.sgd.db)

bb <- prora::get_UniprotID_from_fasta_header(ungroup(merged$merged),idcolumn = "protein_Id")
bb <- prora::map_ids_uniprot(bb)
ranklist <- bb$statistic
names(ranklist) <- bb$P_ENTREZGENEID
res <- clusterProfiler::gseGO(sort(ranklist, decreasing = TRUE), OrgDb = org.Sc.sgd.db, ont = "ALL")
```

```{r fig.cap="ridgeplot", eval = evalAll}
ridgeplot( res )
```

```{r fig.cap = "Dotplot", eval = evalAll}
dotplot(res , showCategory = 30)
```

```{r fig.cap="Upset Plot", eval = evalAll}
enrichplot::upsetplot(res)
```
