---
title: "Contrast Facades with Two-Factor Subgroup Design"
author: "Witold E. Wolski"
date: "`r Sys.Date()`"
output: html_document
vignette: >
  %\VignetteIndexEntry{Contrast Facades with Two-Factor Subgroup Design}
  %\VignetteEncoding{UTF-8}
  %\VignetteEngine{knitr::rmarkdown}
editor_options:
  chunk_output_type: console
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE, message = FALSE, warning = FALSE)
library(prolfqua)
library(dplyr)
library(ggplot2)
```

# Purpose

This vignette demonstrates the same `build_contrast_analysis()` facade workflow
on a dataset with two factors (e.g. treatment and genotype), yielding four
subgroups: `C_WT`, `T_WT`, `C_KO`, `T_KO`. We define contrasts:

- `T_C_gv_WT = T_WT - C_WT`
- `T_C_gv_KO = T_KO - C_KO`
- `WT_KO_comp = T_C_gv_WT - T_C_gv_KO`

We start from simulated two-factor peptide data, derive a single subgroup factor,
and run both protein-input and peptide-input facades with the same contrast
specification. See `ContrastFacades.Rmd` for the one-factor introduction.

# Simulate two-factor data and encode subgroups

```{r simulate_two_factor_subgroups}
dd <- sim_lfq_data_2Factor_config(
  Nprot = 80,
  with_missing = TRUE,
  weight_missing = 1,
  PEPTIDE = TRUE,
  seed = 7
)

cfg_subgroup <- dd$config$clone(deep = TRUE)
data_subgroup <- dd$data |>
  dplyr::mutate(
    Condition = dplyr::recode(Treatment, A = "C", B = "T"),
    Genotype = dplyr::recode(Background, Z = "WT", X = "KO"),
    Subgroup = paste(Condition, Genotype, sep = "_")
  )

cfg_subgroup$factors <- c(Subgroup = "Subgroup")
cfg_subgroup$factorDepth <- 1

lfq_peptide_2f <- LFQData$new(data_subgroup, cfg_subgroup)
lfq_peptide_2f <- lfq_peptide_2f$get_Transformer()$log2()$lfq

lfq_protein_2f <- LFQDataAggregator$new(lfq_peptide_2f, "protein")$medpolish()

contrasts_2f <- c(
  T_C_gv_WT = "SubgroupT_WT - SubgroupC_WT",
  T_C_gv_KO = "SubgroupT_KO - SubgroupC_KO",
  WT_KO_comp = "T_C_gv_WT - T_C_gv_KO"
)
```

# Protein-input facades

```{r facades_protein_two_factor}
fa_lm_2f <- build_contrast_analysis(
  lfq_protein_2f,
  "~ Subgroup",
  contrasts_2f,
  method = "lm"
)

fa_limma_2f <- build_contrast_analysis(
  lfq_protein_2f,
  "~ Subgroup",
  contrasts_2f,
  method = "limma"
)

fa_lm_missing_2f <- build_contrast_analysis(
  lfq_protein_2f,
  "~ Subgroup",
  contrasts_2f,
  method = "lm_missing"
)

fa_deqms_2f <- build_contrast_analysis(
  lfq_protein_2f,
  "~ Subgroup",
  contrasts_2f,
  method = "deqms"
)

fa_firth_protein_2f <- build_contrast_analysis(
  lfq_protein_2f,
  "~ Subgroup",
  contrasts_2f,
  method = "firth"
)
```

```{r collect_protein_two_factor}
results_protein_2f <- bind_rows(
  fa_lm_2f$get_contrasts(),
  fa_limma_2f$get_contrasts(),
  fa_lm_missing_2f$get_contrasts(),
  fa_deqms_2f$get_contrasts(),
  fa_firth_protein_2f$get_contrasts()
)

results_protein_2f |>
  dplyr::count(facade, contrast, name = "n_results")
```

# Peptide-input facades

```{r facades_peptide_two_factor}
fa_lmer_2f <- build_contrast_analysis(
  lfq_peptide_2f,
  "~ Subgroup + (1 | peptide_Id) + (1 | sampleName)",
  contrasts_2f,
  method = "lmer"
)

fa_ropeca_2f <- build_contrast_analysis(
  lfq_peptide_2f,
  "~ Subgroup",
  contrasts_2f,
  method = "ropeca"
)

fa_firth_peptide_2f <- build_contrast_analysis(
  lfq_peptide_2f,
  "~ Subgroup",
  contrasts_2f,
  method = "firth"
)
```

```{r collect_peptide_two_factor}
results_peptide_2f <- bind_rows(
  fa_lmer_2f$get_contrasts(),
  fa_ropeca_2f$get_contrasts(),
  fa_firth_peptide_2f$get_contrasts()
)

results_peptide_2f |>
  dplyr::count(facade, contrast, name = "n_results")
```

# Volcano plots

```{r volcano_protein_2f, fig.height=8, fig.width=10, fig.cap="Protein-input facades, subgroup contrasts."}
results_protein_2f <- results_protein_2f |>
  dplyr::mutate(significant = FDR < 0.1 & abs(diff) > 0.5)

ggplot(results_protein_2f, aes(x = diff, y = -log10(p.value), color = significant)) +
  geom_point(alpha = 0.6, size = 1.2) +
  facet_grid(contrast ~ facade, scales = "free_y") +
  geom_vline(xintercept = c(-0.5, 0.5), linetype = "dashed", color = "grey60") +
  geom_hline(yintercept = -log10(0.1), linetype = "dashed", color = "grey60") +
  scale_color_manual(values = c(`TRUE` = "firebrick", `FALSE` = "grey70")) +
  labs(x = "log2 fold change", y = "-log10(p.value)", color = "FDR < 0.1 & |diff| > 0.5") +
  theme_minimal(base_size = 12)
```

```{r volcano_peptide_2f, fig.height=6, fig.width=9, fig.cap="Peptide-input facades, subgroup contrasts."}
results_peptide_2f <- results_peptide_2f |>
  dplyr::mutate(significant = FDR < 0.1 & abs(diff) > 0.5)

ggplot(results_peptide_2f, aes(x = diff, y = -log10(p.value), color = significant)) +
  geom_point(alpha = 0.6, size = 1.2) +
  facet_grid(contrast ~ facade, scales = "free_y") +
  geom_vline(xintercept = c(-0.5, 0.5), linetype = "dashed", color = "grey60") +
  geom_hline(yintercept = -log10(0.1), linetype = "dashed", color = "grey60") +
  scale_color_manual(values = c(`TRUE` = "firebrick", `FALSE` = "grey70")) +
  labs(x = "log2 fold change", y = "-log10(p.value)", color = "FDR < 0.1 & |diff| > 0.5") +
  theme_minimal(base_size = 12)
```
