---
title: "Comparing Contrast Facades"
author: "Witold E. Wolski"
date: "`r Sys.Date()`"
output: html_document
vignette: >
  %\VignetteIndexEntry{Comparing Contrast Facades}
  %\VignetteEncoding{UTF-8}
  %\VignetteEngine{knitr::rmarkdown}
editor_options:
  chunk_output_type: console
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE, message = FALSE, warning = FALSE)
library(prolfqua)
library(dplyr)
library(ggplot2)
```

# Purpose

`build_contrast_analysis()` provides a common front-end for several contrast
backends:

- `lm`
- `limma`
- `lmer`
- `ropeca`
- `lm_missing`
- `deqms`
- `firth`

All of them expose the same basic interface: `get_contrasts()`,
`get_Plotter()`, and `to_wide()`. The important difference is the required data
level:

- `lm`, `limma`, `lm_missing`, and `deqms` require aggregated protein-level data
- `lmer` and `ropeca` require lower-level measurements such as peptides nested
  within proteins
- `firth` can be used with either aggregated protein-level data or nested
  peptide-level data

This vignette starts from one simulated peptide-level experiment, aggregates it
to protein level, and then demonstrates both families of facades separately.

# Simulate one experiment

```{r simulate}
istar <- sim_lfq_data_peptide_config(Nprot = 80, seed = 42)
istar$config <- old2new(istar$config)

lfq_peptide <- LFQData$new(istar$data, istar$config)
lfq_peptide <- lfq_peptide$get_Transformer()$log2()$lfq

aggregator <- LFQDataAggregator$new(lfq_peptide, "protein")
lfq_protein <- aggregator$medpolish()

lfq_peptide$config$hierarchy_keys()
lfq_protein$config$hierarchy_keys()
lfq_protein$config$nr_children
```

The aggregation step produces protein-level intensities while keeping count
metadata in the `LFQData` object. The DEqMS facade uses
`lfq_protein$config$nr_children` directly, so no extra count table has to be
passed around.

# Define one contrast

```{r contrasts}
contrasts <- c("A_vs_Ctrl" = "group_A - group_Ctrl")
```

Using one contrast keeps the comparisons easy to read while still showing how
the different backends behave.

# Protein-level facades

The following facades require aggregated input, which in practice means
`lfqdata$subject_Id()` must match `lfqdata$config$hierarchy_keys()`.

```{r facades_protein}
fa_lm <- build_contrast_analysis(
  lfq_protein,
  "~ group_",
  contrasts,
  method = "lm"
)

fa_limma <- build_contrast_analysis(
  lfq_protein,
  "~ group_",
  contrasts,
  method = "limma"
)

fa_lm_missing <- build_contrast_analysis(
  lfq_protein,
  "~ group_",
  contrasts,
  method = "lm_missing"
)

fa_deqms <- build_contrast_analysis(
  lfq_protein,
  "~ group_",
  contrasts,
  method = "deqms"
)

fa_firth_protein <- build_contrast_analysis(
  lfq_protein,
  "~ group_",
  contrasts,
  method = "firth"
)
```

Because all protein-level facades share the same interface and the same unit of
analysis, their outputs can be combined directly.

```{r collect_protein}
results_protein <- bind_rows(
  fa_lm$get_contrasts(),
  fa_limma$get_contrasts(),
  fa_lm_missing$get_contrasts(),
  fa_deqms$get_contrasts(),
  fa_firth_protein$get_contrasts()
) |>
  dplyr::select(dplyr::any_of(c(
    "facade", "modelName", "protein_Id", "contrast", "avgAbd", "diff", "FDR",
    "statistic", "std.error", "df", "p.value", "conf.low", "conf.high",
    "sigma"
  ))) |>
  dplyr::mutate(
    significant = FDR < 0.1 & abs(diff) > 0.5
  )

results_protein |>
  dplyr::count(facade, name = "n_results")
```

For facades that combine several underlying result types, such as
`lm_missing`, the `modelName` column still tells you where individual rows came
from.

```{r collect_protein_model_names}
results_protein |>
  dplyr::count(facade, modelName, name = "n_results")
```

# Protein-level volcano comparison

```{r volcano_protein, fig.height=8, fig.width=10, fig.cap="Volcano plots for the protein-level facades. Each panel shows the same contrast analysed by a different backend."}
ggplot(results_protein, aes(x = diff, y = -log10(p.value), color = significant)) +
  geom_point(alpha = 0.6, size = 1.2) +
  facet_wrap(~ facade, scales = "free_y") +
  geom_vline(xintercept = c(-0.5, 0.5), linetype = "dashed", color = "grey60") +
  geom_hline(yintercept = -log10(0.1), linetype = "dashed", color = "grey60") +
  scale_color_manual(values = c(`TRUE` = "firebrick", `FALSE` = "grey70")) +
  labs(
    x = "log2 fold change",
    y = "-log10(p.value)",
    color = "FDR < 0.1\nand |diff| > 0.5"
  ) +
  theme_minimal(base_size = 12)
```

# Looking at the strongest protein-level hits

```{r top_hits_protein}
results_protein |>
  dplyr::group_by(facade) |>
  dplyr::slice_min(order_by = p.value, n = 5, with_ties = FALSE) |>
  dplyr::ungroup() |>
  dplyr::select(facade, modelName, protein_Id, diff, p.value, FDR)
```

# Peptide-level facades

The mixed-effects `lmer` facade and `ropeca` require lower-level measurements
below the analysis subject. The `firth` facade can also operate directly on the
peptide-level `LFQData`, where it models missingness instead of intensity.

```{r facades_peptide}
fa_lmer <- build_contrast_analysis(
  lfq_peptide,
  "~ group_ + (1 | peptide_Id) + (1 | sampleName)",
  contrasts,
  method = "lmer"
)

fa_ropeca <- build_contrast_analysis(
  lfq_peptide,
  "~ group_",
  contrasts,
  method = "ropeca"
)

fa_firth_peptide <- build_contrast_analysis(
  lfq_peptide,
  "~ group_",
  contrasts,
  method = "firth"
)
```

`ropeca` aggregates peptide evidence back to proteins, whereas `lmer` models the
nested peptide structure directly before reporting protein-level contrasts.
For peptide-level `firth`, proteins with exactly one peptide are fitted without
an added peptide term, while proteins with multiple peptides are fitted with the
lowest hierarchy key added internally.

```{r collect_peptide}
results_peptide <- bind_rows(
  fa_lmer$get_contrasts(),
  fa_ropeca$get_contrasts(),
  fa_firth_peptide$get_contrasts()
) |>
  dplyr::select(dplyr::any_of(c(
    "facade", "modelName", "protein_Id", "contrast", "avgAbd", "diff", "FDR",
    "statistic", "std.error", "df", "p.value", "conf.low", "conf.high",
    "sigma"
  ))) |>
  dplyr::mutate(
    significant = FDR < 0.1 & abs(diff) > 0.5
  )

results_peptide |>
  dplyr::count(facade, name = "n_results")
```

```{r volcano_peptide, fig.height=5, fig.width=9, fig.cap="Volcano plots for the peptide-level facades."}
ggplot(results_peptide, aes(x = diff, y = -log10(p.value), color = significant)) +
  geom_point(alpha = 0.6, size = 1.2) +
  facet_wrap(~ facade, scales = "free_y") +
  geom_vline(xintercept = c(-0.5, 0.5), linetype = "dashed", color = "grey60") +
  geom_hline(yintercept = -log10(0.1), linetype = "dashed", color = "grey60") +
  scale_color_manual(values = c(`TRUE` = "firebrick", `FALSE` = "grey70")) +
  labs(
    x = "log2 fold change",
    y = "-log10(p.value)",
    color = "FDR < 0.1\nand |diff| > 0.5"
  ) +
  theme_minimal(base_size = 12)
```

# Remarks

The facades make it easy to benchmark alternative contrast backends without
rewriting the analysis pipeline:

- the protein-level facades now enforce aggregation before modelling
- the peptide-level facades now explicitly require lower-level hierarchy below
  the analysis subject
- the shared facade API still makes it straightforward to compare methods once
  the data level is chosen consistently

The results are comparable at the API level, but the comparison is only
meaningful when methods using the same biological unit are plotted together.
