---
title: "Contrasts from factors"
author: "Witold E. Wolski"
date: "`r Sys.Date()`"
output:
  html_document: default
  pdf_document: default
vignette: |
  %\VignetteIndexEntry{Contrasts from factors}
  %\VignetteEncoding{UTF-8}
  %\VignetteEngine{knitr::rmarkdown}
editor_options:
  chunk_output_type: console
bibliography:
  - prolfqua.bib
---


```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE, message = FALSE, warning = FALSE)
```

# Purpose

When working with factorial experimental designs (e.g. two conditions crossed with three time points), specifying all the relevant contrasts by hand is tedious and error-prone. The `generate_contrasts` family of functions in prolfqua automates this by generating contrast specifications from factor levels. These contrast strings can then be passed directly to the `Contrasts` class for statistical testing.

This vignette demonstrates how to:

- Generate main effect, level-specific, and interaction contrasts for a two-factor design
- Use `annotation_add_contrasts` to produce a combined annotation and contrast table ready for analysis

# Group labelling convention

All contrast generation functions assume that the group levels in the fitted model follow the naming convention `G_<primary>_<secondary>`, which is produced by `group_label()`:

```{r group_label}
library(prolfqua)

group_label("MI", "T0")
group_label("MINOCA", "T300")
```

This means that before fitting a model, the data must contain a grouping column with levels in this format. The `annotation_add_contrasts` function creates such a column automatically using `tidyr::unite`.

# Building contrasts step by step

Consider a two-factor design with disease type (MI, MINOCA) and time point (T0, T150, T300):

```{r define_levels}
primary_levels <- c("MI", "MINOCA")
secondary_levels <- c("T0", "T150", "T300")
```

## Main effect contrasts

Main effects average across all levels of the secondary factor. For example, the main effect of MINOCA vs MI is the average difference across all time points:

```{r main_effects}
me <- main_effect_contrasts(primary_levels, secondary_levels)
data.frame(ContrastName = names(me), Contrast = unlist(me))
```

Swapping the roles of primary and secondary gives main effects for time points averaged across disease types:

```{r main_effects_swapped}
me2 <- main_effect_contrasts(secondary_levels, primary_levels)
data.frame(ContrastName = names(me2), Contrast = unlist(me2))
```

## Level-specific contrasts

These compare primary factor levels at each individual level of the secondary factor:

```{r level_specific}
ls <- level_specific_contrasts(primary_levels, secondary_levels)
data.frame(ContrastName = names(ls), Contrast = unlist(ls))
```

## Interaction contrasts

Interaction contrasts test whether the difference between primary levels changes across secondary levels (difference of differences):

```{r interactions}
ic <- interaction_contrasts(primary_levels, secondary_levels)
data.frame(ContrastName = names(ic), Contrast = unlist(ic))
```

## Single-factor contrasts

For a one-factor design, `generate_contrasts_for_factor` generates all pairwise comparisons:

```{r single_factor}
group_levels <- c("CondA", "CondB", "CondC")
sf <- generate_contrasts_for_factor(group_levels)
data.frame(ContrastName = names(sf), Contrast = unlist(sf))
```

# Generating all contrasts at once

`generate_contrasts` combines main effects, level-specific, and interaction contrasts into a single data frame:

```{r generate_all}
all_contrasts <- generate_contrasts(primary_levels, secondary_levels)
knitr::kable(all_contrasts, row.names = FALSE)
```

To exclude interaction contrasts:

```{r generate_no_interactions}
no_int <- generate_contrasts(primary_levels, secondary_levels, interactions = FALSE)
knitr::kable(no_int, row.names = FALSE)
```

# Working with annotation tables

In a typical prolfquapp workflow, you start with a sample annotation table that has columns for the two factors. `annotation_add_contrasts` creates a united Group column, generates all contrasts, and binds them alongside the annotation:

```{r annotation_example}
# x5463yzwer453bbb is a bundled example annotation table
# with factor_A (MI / MINOCA) and factor_B (T0 / T150 / T300)
data("x5463yzwer453bbb", package = "prolfqua")
head(x5463yzwer453bbb[, c("Name", "Group", "factor_A", "factor_B")])
```

```{r annotation_add_contrasts}
result <- annotation_add_contrasts(
  x5463yzwer453bbb,
  primary_col = "factor_A",
  secondary_col = "factor_B",
  prefix = "primary"
)

# The annotation with united Group and contrast columns
knitr::kable(head(result$annot[, c("Name", "Group", "ContrastName", "Contrast")], 10),
             row.names = FALSE)

# Suggested output file name
result$name
```

Swapping primary and secondary factors gives contrasts from the other perspective:

```{r annotation_swapped}
result2 <- annotation_add_contrasts(
  x5463yzwer453bbb,
  primary_col = "factor_B",
  secondary_col = "factor_A",
  prefix = "secondary"
)
knitr::kable(head(result2$annot[, c("Name", "Group", "ContrastName", "Contrast")], 10),
             row.names = FALSE)
```

# Session Info

```{r session_info}
sessionInfo()
```
