---
title: "Contrast Facades with Parallel Designs"
author: "Witold E. Wolski"
date: "`r Sys.Date()`"
output: html_document
vignette: >
  %\VignetteIndexEntry{Contrast Facades with Parallel Designs}
  %\VignetteEncoding{UTF-8}
  %\VignetteEngine{knitr::rmarkdown}
editor_options:
  chunk_output_type: console
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE, message = FALSE, warning = FALSE)
library(prolfqua)
library(dplyr)
library(ggplot2)
```

# Purpose

`build_contrast_analysis()` provides a common front-end for several contrast
backends:

- `lm` — ordinary least-squares linear model fitted per protein, Wald contrasts.
- `rfit` — rank-based robust regression (`Rfit::rfit`); like `lm` but resistant to outliers in the response.
- `limma` — linear model with empirical-Bayes moderated variance shrinkage across proteins.
- `limma_impute` — `limma` with LOD imputation and borrowed per-protein variance for groups with no observations.
- `limpa` — vooma precision-weighted model on DPC-aggregated proteins (requires `AggregateLimpa` for DPC-based aggregation with standard errors).
- `lmer` — linear mixed-effects model on nested peptide/precursor data, reporting protein-level contrasts.
- `ropeca` — peptide-level fold changes aggregated to protein via ROPECA median + beta-distribution p-values.
- `lm_missing` — `lm` merged with group-mean substitution for proteins `lm` cannot estimate (deprecated; prefer `lm_impute`).
- `lm_impute` — `lm` that refits failed/singular proteins with LOD imputation and borrowed variance.
- `deqms` — `limma` plus DEqMS count-dependent variance moderation using peptide counts per protein.
- `firth` — Firth bias-reduced logistic regression on presence/absence (missingness) of aggregated proteins.
- `firth_nested` — Firth logistic regression on peptide-level missingness, reporting protein-level contrasts.

All of them expose the same basic interface: `get_contrasts()`,
`get_Plotter()`, and `to_wide()`. All examples below return protein-level
contrasts. The important difference is the required input level:

- `lm`, `rfit`, `limma`, `limma_impute`, `lm_missing`, `lm_impute`, and `deqms` require aggregated protein-level data
- `limpa` requires protein-level data from `AggregateLimpa` (which provides standard errors and observation counts for vooma precision weighting)
- `lmer` and `ropeca` require lower-level measurements such as peptides nested
  within proteins, but still report protein-level contrasts
- `firth` can be used with either aggregated protein-level data or nested
  peptide-level data and still reports protein-level contrasts

This vignette starts from one simulated peptide-level experiment, aggregates it
to protein level, and then demonstrates both families of facades separately.

# Simulate one experiment

```{r simulate}
options(prolfqua.vectorize = TRUE)

istar <- sim_lfq_data_peptide_config(Nprot = 80, seed = 42)


lfq_peptide <- LFQData$new(istar$data, istar$config)
lfq_peptide <- lfq_peptide$get_Transformer()$log2()$lfq

lfq_protein <- lfq_peptide$get_Aggregator()$aggregate()

lfq_peptide$hierarchy_keys()
lfq_protein$hierarchy_keys()
lfq_protein$nr_children_col()
```

The aggregation step produces protein-level intensities while keeping count
metadata in the `LFQData` object. The DEqMS facade uses
`lfq_protein$nr_children_col()` directly, so no extra count table has to be
passed around.

# Define contrasts

```{r contrasts}
contrasts <- c(
  "A_vs_Ctrl" = "group_A - group_Ctrl",
  "B_vs_Ctrl" = "group_B - group_Ctrl"
)
```

Two contrasts let us see how each backend handles multiple comparisons and
how FDR correction propagates across contrasts.

# Protein-input facades

The following facades require aggregated input, which in practice means
`lfqdata$subject_id()` must match `lfqdata$hierarchy_keys()`. `firth`
is included here on purpose because it can be fitted directly on aggregated
protein input.

```{r facades_protein}
fa_lm <- build_contrast_analysis(
  lfq_protein,
  "~ group_",
  contrasts,
  method = "lm"
)

fa_rfit <- build_contrast_analysis(
  lfq_protein,
  "~ group_",
  contrasts,
  method = "rfit"
)

fa_limma <- build_contrast_analysis(
  lfq_protein,
  "~ group_",
  contrasts,
  method = "limma"
)

fa_limma_impute <- build_contrast_analysis(
  lfq_protein,
  "~ group_",
  contrasts,
  method = "limma_impute"
)

fa_lm_missing <- build_contrast_analysis(
  lfq_protein,
  "~ group_",
  contrasts,
  method = "lm_missing"
)

fa_deqms <- build_contrast_analysis(
  lfq_protein,
  "~ group_",
  contrasts,
  method = "deqms"
)

fa_lm_impute <- build_contrast_analysis(
  lfq_protein,
  "~ group_",
  contrasts,
  method = "lm_impute"
)

fa_firth_protein <- build_contrast_analysis(
  lfq_protein,
  "~ group_",
  contrasts,
  method = "firth"
)
```

### limpa facade (DPC-based aggregation)

The `limpa` facade is different from the other protein-level facades: it requires
its own aggregation step via `AggregateLimpa`, which uses limpa's Detection
Probability Curve (DPC) to aggregate peptides to proteins while producing
per-protein, per-sample standard errors. These SEs feed into a bivariate vooma
variance model for precision weighting.

```{r facade_limpa, eval=requireNamespace("limpa", quietly = TRUE)}
lfq_limpa <- AggregateLimpa$new(lfq_peptide, "protein")$aggregate()

fa_limpa <- build_contrast_analysis(
  lfq_limpa,
  "~ group_",
  contrasts,
  method = "limpa"
)
```

Because all protein-input facades share the same interface and report
protein-level contrasts, their outputs can be combined directly.

```{r collect_protein}
# Proteins missing in the baseline lm facade (used to flag rescued proteins)
lm_missing_ids <- fa_lm$get_missing() |>
  dplyr::select(protein_Id, contrast) |>
  dplyr::mutate(rescued = TRUE)

results_limpa <- if (requireNamespace("limpa", quietly = TRUE)) {
  fa_limpa$get_contrasts()
} else {
  data.frame()
}

results_protein <- bind_rows(
  fa_lm$get_contrasts(),
  fa_rfit$get_contrasts(),
  fa_limma$get_contrasts(),
  fa_limma_impute$get_contrasts(),
  fa_lm_missing$get_contrasts(),
  fa_lm_impute$get_contrasts(),
  fa_deqms$get_contrasts(),
  fa_firth_protein$get_contrasts(),
  results_limpa
) |>
  dplyr::select(dplyr::any_of(c(
    "modelName", "estimate_type", "protein_Id", "contrast", "avgAbd", "diff", "FDR",
    "statistic", "std.error", "df", "p.value", "conf.low", "conf.high",
    "sigma"
  ))) |>
  dplyr::left_join(lm_missing_ids, by = c("protein_Id", "contrast")) |>
  dplyr::mutate(
    rescued = dplyr::coalesce(rescued, FALSE),
    significant = FDR < 0.1 & abs(diff) > 0.5
  )

results_protein |>
  dplyr::count(modelName, name = "n_results")
```

The `modelName` column is the facade key. For facades that combine several
underlying result types, such as `lm_missing`, the `estimate_type` column tells
you how each row was produced (`observed`, `lod_imputed`, or `missing_fallback`).

```{r collect_protein_model_names}
results_protein |>
  dplyr::count(modelName, contrast, estimate_type, name = "n_results")
```

# Protein-level volcano comparison

## Standard facades

```{r volcano_standard, fig.height=8, fig.width=12, fig.cap="Volcano plots for the standard protein-level facades (lm, rfit, limma, deqms)."}
standard_facades <- c("lm", "rfit", "limma", "deqms")
results_standard <- results_protein |>
  dplyr::filter(modelName %in% standard_facades)

ggplot(results_standard, aes(x = diff, y = -log10(p.value), color = significant)) +
  geom_point(alpha = 0.6, size = 1.5) +
  facet_grid(contrast ~ modelName, scales = "free_y") +
  geom_vline(xintercept = c(-0.5, 0.5), linetype = "dashed", color = "grey60") +
  geom_hline(yintercept = -log10(0.1), linetype = "dashed", color = "grey60") +
  scale_color_manual(values = c(`TRUE` = "firebrick", `FALSE` = "grey70")) +
  labs(x = "log2 fold change", y = "-log10(p.value)", color = "FDR < 0.1\nand |diff| > 0.5") +
  theme_minimal(base_size = 12)
```

## Imputation and missingness facades

Rescued proteins (missing in plain `lm`) are shown as large red diamonds
so they stand out clearly. This group includes the LOD imputation facades
(`lm_missing`, `lm_impute`, `limma_impute`) and the firth logistic
regression facade which models missingness directly.

```{r volcano_impute, fig.height=8, fig.width=14, fig.cap="Volcano plots for the imputation and missingness facades. Large red diamonds mark proteins rescued (missing in plain lm)."}
impute_facades <- c("lm_missing", "lm_impute", "limma_impute", "firth")
results_impute <- results_protein |>
  dplyr::filter(modelName %in% impute_facades) |>
  dplyr::mutate(modelName = factor(modelName, levels = impute_facades))

ggplot(results_impute, aes(x = diff, y = -log10(p.value))) +
  geom_point(data = dplyr::filter(results_impute, !rescued),
             aes(color = significant), alpha = 0.5, size = 1.2) +
  geom_point(data = dplyr::filter(results_impute, rescued),
             color = "red", shape = 18, size = 5, alpha = 1) +
  facet_grid(contrast ~ modelName, scales = "free_y") +
  geom_vline(xintercept = c(-0.5, 0.5), linetype = "dashed", color = "grey60") +
  geom_hline(yintercept = -log10(0.1), linetype = "dashed", color = "grey60") +
  scale_color_manual(values = c(`TRUE` = "firebrick", `FALSE` = "grey70")) +
  labs(x = "log2 fold change", y = "-log10(p.value)", color = "FDR < 0.1\nand |diff| > 0.5") +
  theme_minimal(base_size = 12)
```

# Looking at the strongest protein-level hits

```{r top_hits_protein}
results_protein |>
  dplyr::group_by(modelName, contrast) |>
  dplyr::slice_min(order_by = p.value, n = 5, with_ties = FALSE) |>
  dplyr::ungroup() |>
  dplyr::select(modelName, estimate_type, contrast, protein_Id, diff, p.value, FDR)
```

# Proteins that could not be estimated

Every facade has a `get_missing()` method that returns the protein × contrast
pairs present in the input data but absent from `get_contrasts()`. This makes
it easy to see which proteins each method fails on and to compare coverage.

```{r missing_protein_facades}
missing_limpa <- if (requireNamespace("limpa", quietly = TRUE)) {
  fa_limpa$get_missing() |> dplyr::mutate(facade = "limpa")
} else {
  data.frame()
}

missing_all <- dplyr::bind_rows(
  fa_lm$get_missing() |> dplyr::mutate(facade = "lm"),
  fa_limma$get_missing() |> dplyr::mutate(facade = "limma"),
  fa_limma_impute$get_missing() |> dplyr::mutate(facade = "limma_impute"),
  fa_lm_missing$get_missing() |> dplyr::mutate(facade = "lm_missing"),
  fa_lm_impute$get_missing() |> dplyr::mutate(facade = "lm_impute"),
  fa_deqms$get_missing() |> dplyr::mutate(facade = "deqms"),
  fa_firth_protein$get_missing() |> dplyr::mutate(facade = "firth"),
  missing_limpa
)

missing_all |>
  dplyr::count(facade, contrast, name = "n_missing") |>
  knitr::kable(caption = "Number of missing protein × contrast pairs per facade")
```

## Per-sample intensities of the missing proteins

```{r missing_intensities}
missing_proteins <- unique(missing_all$protein_Id)

if (length(missing_proteins) > 0) {
  lfq_protein$data_long() |>
    dplyr::filter(protein_Id %in% missing_proteins) |>
    dplyr::select(protein_Id, sampleName,
                  !!rlang::sym(lfq_protein$response())) |>
    tidyr::pivot_wider(names_from = sampleName,
                       values_from = !!rlang::sym(lfq_protein$response())) |>
    knitr::kable(digits = 2, caption = "Per-sample intensities of proteins that could not be estimated")
}
```

The missing cells (NA) explain why these proteins cannot be estimated — they
lack observations in one or more groups. The `lm_missing` facade fills in
these gaps via group-mean imputation, while `lm_impute` and `limma_impute`
re-fit after imputing individual values with the LOD and borrowing covariance
from successful fits. These should have fewer missing proteins than plain
`lm` or `limma`.

## Estimates for the missing proteins from `lm_missing` and `lm_impute`

For proteins that plain `lm` could not estimate, the imputation-based facades
can still produce contrast results. The table below shows these rescued
estimates side by side.

```{r rescued_estimates}
lm_missing_proteins <- fa_lm$get_missing()$protein_Id |> unique()

if (length(lm_missing_proteins) > 0) {
  rescued <- results_protein |>
    dplyr::filter(
      protein_Id %in% lm_missing_proteins,
      modelName %in% c("lm_missing", "lm_impute", "limma_impute", "limpa")
    ) |>
    dplyr::arrange(protein_Id, contrast, modelName)

  rescued |>
    knitr::kable(
      digits = 3,
      caption = "Contrast estimates from lm_missing, lm_impute, and limma_impute for proteins that plain lm could not estimate"
    )
} else {
  cat("All proteins were estimable by plain lm — no rescued estimates to show.")
}
```

# Peptide-input facades

The peptide-input facades take peptide/precursor-level `LFQData` and return
protein-level contrasts. They are defined in `R/ContrastsChildToParentFacades.R`
and use the `*_nested` method keys: `lmer_nested`, `ropeca_nested`,
`firth_nested`, and `limpa_nested`. `firth_nested` is the peptide-input
counterpart of the protein-input `firth` shown above.

```{r facades_peptide}
fa_lmer <- build_contrast_analysis(
  lfq_peptide,
  "~ group_ + (1 | peptide_Id) + (1 | sampleName)",
  contrasts,
  method = "lmer_nested"
)

fa_ropeca <- build_contrast_analysis(
  lfq_peptide,
  "~ group_",
  contrasts,
  method = "ropeca_nested"
)

fa_firth_peptide <- build_contrast_analysis(
  lfq_peptide,
  "~ group_",
  contrasts,
  method = "firth_nested"
)
```

`ropeca_nested` aggregates peptide evidence back to proteins, whereas
`lmer_nested` models the nested peptide structure directly before reporting
protein-level contrasts. `firth_nested` also reports protein-level contrasts:
proteins with exactly one peptide are fitted without an added peptide term,
while proteins with multiple peptides are fitted with the lowest hierarchy
key added internally.

```{r collect_peptide}
results_peptide <- bind_rows(
  fa_lmer$get_contrasts(),
  fa_ropeca$get_contrasts(),
  fa_firth_peptide$get_contrasts()
) |>
  dplyr::select(dplyr::any_of(c(
    "modelName", "estimate_type", "protein_Id", "contrast", "avgAbd", "diff", "FDR",
    "statistic", "std.error", "df", "p.value", "conf.low", "conf.high",
    "sigma"
  ))) |>
  dplyr::mutate(
    significant = FDR < 0.1 & abs(diff) > 0.5
  )

results_peptide |>
  dplyr::count(modelName, name = "n_results")
```

```{r volcano_peptide, fig.height=8, fig.width=10, fig.cap="Volcano plots for the peptide-level facades. Rows are contrasts, columns are backends."}
ggplot(results_peptide, aes(x = diff, y = -log10(p.value), color = significant)) +
  geom_point(alpha = 0.6, size = 1.2) +
  facet_grid(contrast ~ modelName, scales = "free_y") +
  geom_vline(xintercept = c(-0.5, 0.5), linetype = "dashed", color = "grey60") +
  geom_hline(yintercept = -log10(0.1), linetype = "dashed", color = "grey60") +
  scale_color_manual(values = c(`TRUE` = "firebrick", `FALSE` = "grey70")) +
  labs(
    x = "log2 fold change",
    y = "-log10(p.value)",
    color = "FDR < 0.1\nand |diff| > 0.5"
  ) +
  theme_minimal(base_size = 12)
```

# Remarks

The facades make it easy to benchmark alternative contrast backends without
rewriting the analysis pipeline:

- the protein-level facades now enforce aggregation before modelling
- the `limpa` facade uses its own DPC-based aggregation (`AggregateLimpa`)
  which produces standard errors propagated into vooma precision weights
- the peptide-level facades now explicitly require lower-level hierarchy below
  the analysis subject
- the shared facade API still makes it straightforward to compare methods once
  the data level is chosen consistently

The results are comparable at the API level, but the comparison is only
meaningful when methods using the same biological unit are plotted together.

# Session Info

```{r session_info}
sessionInfo()
```
