---
title: "Creating prolfqua Configurations"
author: "Witold E. Wolski"
date: "`r Sys.Date()`"
output: html_document
vignette: >
  %\VignetteIndexEntry{Creating prolfqua Configurations} 
  %\VignetteEncoding{UTF-8}
  %\VignetteEngine{knitr::rmarkdown}
bibliography:
  - prolfqua.bib
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE, message = FALSE, warning = FALSE)
```

# Creating a configuration for a file in long data format.

Given for example a Peptide Quantification Report generated by Spectronaut (a table in long format), we demonstrate how to create a Configuration that is required to use it with prolfqua. To do this, an `AnalysisTableAnnotation` has to be configured and some fields (fileName, hierarchy, factors, workingIntensity) need to defined.
The configuration object describes the columns in the long table so that prolfqua functions know which columns to use.

```{r readInData}
conflicted::conflict_prefer("rename", "dplyr")
file <- prolfqua:::.find.package.file("prolfqua","samples/spectronaut_PEPTIDE.xls")
xx <- readr::read_tsv(file)
xx <- xx |> dplyr::rename(EG.TotalQuantity = "EG.TotalQuantity (Settings)")
xx$Isotope.Label <- "light"
```

We create a Table annotation object and start annotating the data we read. Since in this example we eventually want to do more filtering on data quality we will also define the ident_qValue in this `AnalysisTableAnnotation`.

```{r defineAnalysisTable}
atable <- prolfqua::AnalysisTableAnnotation$new()
atable$fileName = "R.FileName"
atable$ident_qValue = "EG.Qvalue"
atable$workIntensity = "EG.TotalQuantity"
atable$isotopeLabel = "Isotope.Label"

```

The columns identifying the measured features, which are proteins, peptides or precursors, are the described using the named list hierarchy. The values of the list are the column names, while the names of your list are arbitrary as long as they are valid R column names (see function `make.names`). 

The list factors, is used to point to the columns containing the factors of your analysis.
Here, we rename the column "R.Condition" to "Marker". In figures and legends generated by prolfqua the name "Marker" will then be used and not "R.Condition". The data.frame can also contain more than one factor. 


```{r defineHierarchyAndFactors}
atable$hierarchy[["PG.ProteinAccessions"]]    <-  "PG.ProteinAccessions"
atable$hierarchy[["EG.PrecursorId"]]    <-  "EG.PrecursorId"
atable$factors[["Marker"]] <- "R.Condition"
```


Lastly, we create an Analysis parameter object, and the Analysis Configuration. The function `setup_analysis`, creates from data frame in long format a data.frame compatible with your configuration. We can now run most of the function in the package using the data and configuration.

```{r mateDataAndConfig}
config <- prolfqua::AnalysisConfiguration$new(atable)
analysis_data <- prolfqua::setup_analysis(xx, config)

prolfqua::summarize_hierarchy(analysis_data, config)

```

Now the analysis_data object is ready to generate the `LFQData` class instance. This object is the start for further analysis.

```{r generateLFQData}
lfqdata <- prolfqua::LFQData$new(analysis_data, config)
```

With this, it is possible for example to use the `get_Summariser` function to visualize and summarise the data efficiently.

```{r doBarPlotPerSample, fig.cap="Shown are how many precursors or proteins are found in each sample", fig.align=5, fig.height=5}
smrz <- lfqdata$get_Summariser()
smrz$plot_hierarchy_counts_sample()
```

The `prolfqua` package is described in [@Wolski2022.06.07.494524].

# Session Info

```{r}
sessionInfo()
```

# References


