---
title: "DEqMS-style Count-Dependent Variance Moderation"
author: "Witold E. Wolski"
date: "`r Sys.Date()`"
output: html_document
vignette: >
  %\VignetteIndexEntry{DEqMS-style Count-Dependent Variance Moderation}
  %\VignetteEncoding{UTF-8}
  %\VignetteEngine{knitr::rmarkdown}
editor_options:
  chunk_output_type: console
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE, message = FALSE, warning = FALSE)
```

# Purpose

This vignette demonstrates `ContrastsModeratedDEqMS`, a count-dependent variance
shrinkage method inspired by the DEqMS approach (Zhu et al., 2020). It is a
drop-in alternative to `ContrastsModerated`:

- **`ContrastsModerated`** (standard): applies a **single global prior variance**
  to all proteins — every protein is shrunk toward the same value.
- **`ContrastsModeratedDEqMS`** (count-dependent): the prior variance **depends on
  the number of peptides/PSMs** used to quantify each protein. Proteins
  quantified from many peptides get less shrinkage (their variance is already
  well estimated); proteins from few peptides get more.

This matters in proteomics because quantification depth varies hugely across
proteins. A protein quantified from 20 peptides has a much better-estimated
variance than one from 2 peptides, and they should not receive the same prior.

The user-facing API is identical to `ContrastsModerated`: same downstream
methods (`get_contrasts`, `get_Plotter`, `to_wide`, `merge_contrasts_results`).

# How it works

1. **Fit LOESS**: `log(sigma^2) ~ log2(peptide_count)` — captures the
   count-dependent variance trend.
2. **Estimate prior df (d0)**: from the residual variance around the LOESS
   curve, using the `trigammaInverse` approach (same math as limma/squeezeVarRob).
3. **Count-specific prior variance**: `s0^2(count)` varies per protein, derived
   from the fitted LOESS curve.
4. **Posterior variance**: Bayesian shrinkage combining observed and prior:
   `var.post = (d0 * s0^2(count) + df * sigma^2) / (d0 + df)`
5. **Moderated t-statistics and p-values** are recomputed with the posterior variance.

# Data preparation

We use simulated protein-level data. The simulation includes a `nr_peptides`
column — the number of peptides per protein — which is exactly the count we need.

```{r data}
library(prolfqua)
library(dplyr)
library(ggplot2)

istar <- sim_lfq_data_protein_config(Nprot = 200, weight_missing = 0.3)
lfqdata <- LFQData$new(istar$data, istar$config)
lfqdata$remove_small_intensities()

lt <- lfqdata$get_Transformer()
transformed <- lt$log2()$robscale()$lfq
transformed$rename_response("transformedIntensity")
```

# Define contrasts

```{r contrasts}
contr_spec <- c(
  "AvsCtrl" = "group_A - group_Ctrl",
  "BvsCtrl" = "group_B - group_Ctrl",
  "AvsB"    = "group_A - group_B"
)
```

# Build the count table

`ContrastsModeratedDEqMS` requires a data.frame mapping each protein to its
peptide count. This is typically extracted from the input data. The constructor
automatically aggregates to one value per protein (taking the max).

```{r count_df}
count_df <- istar$data |>
  select(all_of(c(istar$config$hierarchy_keys_depth(), "nr_peptides"))) |>
  distinct()

# Inspect the distribution of peptide counts
count_per_prot <- count_df |>
  group_by(protein_Id) |>
  summarise(nr_peptides = max(nr_peptides, na.rm = TRUE))

ggplot(count_per_prot, aes(x = nr_peptides)) +
  geom_histogram(binwidth = 1, fill = "steelblue", color = "white") +
  labs(x = "Number of peptides per protein", y = "Count",
       title = "Distribution of peptide counts") +
  theme_minimal()
```

# Standard moderation pipeline

```{r standard}
mod_lm <- build_model(transformed, strategy_lm("transformedIntensity ~ group_"))
contr_lm <- Contrasts$new(mod_lm, contr_spec)
contr_moderated <- ContrastsModerated$new(contr_lm)
res_moderated <- contr_moderated$get_contrasts()
```

# DEqMS moderation pipeline

The only difference: wrap with `ContrastsModeratedDEqMS` instead of
`ContrastsModerated`, and supply `count_df` + `count_column`.

```{r deqms}
contr_deqms <- ContrastsModeratedDEqMS$new(
  contr_lm,
  count_df = count_df,
  count_column = "nr_peptides"
)
res_deqms <- contr_deqms$get_contrasts()
```

# Diagnostic: variance vs. peptide count

This is the core diagnostic plot for the DEqMS approach. We expect `log(sigma^2)`
to decrease with increasing peptide count, and the LOESS curve captures this trend.

```{r diagnostic, fig.cap="Residual variance decreases with peptide count. The LOESS curve (blue) defines the count-specific prior variance."}
# Get the unmoderated results with sigma and count
contr_raw <- contr_lm$get_contrasts()
contr_raw <- inner_join(
  contr_raw,
  count_per_prot,
  by = "protein_Id"
)

# Plot for one contrast
one_contrast <- contr_raw |> filter(contrast == levels(factor(contrast))[1])

ggplot(one_contrast, aes(x = log2(nr_peptides), y = log(sigma^2))) +
  geom_point(alpha = 0.4, size = 2) +
  geom_smooth(method = "loess", span = 0.75, color = "steelblue", se = TRUE) +
  labs(x = "log2(peptide count)",
       y = "log(residual variance)",
       title = paste("Variance vs. peptide count —", one_contrast$contrast[1])) +
  theme_minimal()
```

# Comparing results

## Fold changes are identical

Both decorators wrap the same underlying `Contrasts` object, so fold changes
are identical.

```{r compare_fc}
merged <- inner_join(
  select(res_deqms, protein_Id, contrast, diff_deqms = diff),
  select(res_moderated, protein_Id, contrast, diff_moderated = diff),
  by = c("protein_Id", "contrast")
)

cor(merged$diff_deqms, merged$diff_moderated, use = "complete.obs")
```

## P-values differ (count-dependent shrinkage)

The DEqMS approach produces different p-values because the prior variance
is protein-specific rather than global.

```{r compare_pval}
merged_p <- inner_join(
  select(res_deqms, protein_Id, contrast, p_deqms = p.value),
  select(res_moderated, protein_Id, contrast, p_moderated = p.value),
  by = c("protein_Id", "contrast")
)

cor(-log10(merged_p$p_deqms), -log10(merged_p$p_moderated), use = "complete.obs")
```

```{r plot_pval, fig.cap="P-values: DEqMS vs. standard moderated. Highly correlated but not identical — DEqMS adjusts for peptide count."}
ggplot(merged_p, aes(x = -log10(p_moderated), y = -log10(p_deqms))) +
  geom_point(alpha = 0.3, size = 1.5) +
  geom_abline(slope = 1, intercept = 0, color = "red", linetype = "dashed") +
  labs(x = "-log10(p) standard moderated",
       y = "-log10(p) DEqMS moderated") +
  theme_minimal()
```

# Volcano plots

```{r volcano, fig.cap="Volcano plots from both moderation approaches."}
pl_moderated <- contr_moderated$get_Plotter()
pl_deqms <- contr_deqms$get_Plotter()

gridExtra::grid.arrange(
  pl_moderated$volcano()$FDR + ggtitle("Standard moderated"),
  pl_deqms$volcano()$FDR + ggtitle("DEqMS moderated"),
  ncol = 1
)
```

# Merging with missing value imputation

`ContrastsModeratedDEqMS` works seamlessly with `ContrastsMissing` and
`merge_contrasts_results`.

```{r merge_missing}
mC <- ContrastsMissing$new(lfqdata = transformed, contrasts = contr_spec)
merged_result <- merge_contrasts_results(prefer = contr_deqms, add = mC)

plotter <- merged_result$merged$get_Plotter()
plotter$volcano()$FDR
```

# Wide format export

```{r wide}
wide <- contr_deqms$to_wide()
head(wide)
```

# Inspecting all columns

Use `all = TRUE` to see both the original and moderated columns side by side.

```{r all_columns}
res_all <- contr_deqms$get_contrasts(all = TRUE)
res_all |>
  select(protein_Id, contrast, sigma, diff, statistic,
         moderated.var.prior, moderated.df.prior,
         moderated.var.post, moderated.statistic, moderated.p.value) |>
  head(10)
```

Note that `moderated.var.prior` varies per protein (count-dependent), unlike
in `ContrastsModerated` where it is constant.

# Session Info

```{r}
sessionInfo()
```
