## 1.1. Install R

A Windows/Linux/MacOSX x64 platform R 4 

Make sure you have installed the required packages. Start R session and run the following R code.

```{r}
#"C:\\Users\\wolski\\Documents\\R\\win-library\\4.0"
tmpdir <- tempdir()

oldUserLib <- Sys.getenv("R_LIBS_USER")
oldUserLib

oldLibPaths <- .libPaths()
oldLibPaths

#Overwirte
.libPaths(c(tmpdir,oldLibPaths[[2]]))
Sys.setenv(R_LIBS_USER = tmpdir)

.libPaths()
library(LFQService)
library(remotes)
unlink(tmpdir, recursive = TRUE)

```

```{r}

install.packages('remotes')
# If you want to build the vignettes comment out the next line.
# remotes::install_github('wolski/prolfquaData')
remotes::install_github('wolski/LFQService')

```


```{r}
pkgs <- c('broom', 'bookdown', 'conflicted', 'corrplot', 'dplyr',
  'flextable', 'docopt',
  'ggplot2', 'ggbeeswarm', 'ggfortify', 'glue', 'GGally', 'pheatmap',
  'kableExtra', 'limma', 'lme4', 'lmerTest', 'magrittr', 'multcomp',
  'plotly', 'purrr', 'readxl', 'tidyverse', 'yaml',
  'tidyr', 'writexl')

pkgs <- pkgs[(!pkgs %in% unique(installed.packages()[,'Package']))]
if(length(pkgs) > 0){install.packages(pkgs)}



```

To install the package please execute:






## 2. Running R-scripts

Generate QC report from bat file.
First add `<LFQService_path>/win` to your path variable. Then you can generate a QC report from a maxquant QC by running.


```
lfq_MQ_SampleSizeReport.bat .\data\1296877_QC.zip
```


## 3. Best of code snippets


```
Rscript ~/__checkouts/R/LFQService/inst/run_scripts/lfq_MQ_SampleSizeReport.R --help
Rscript ~/__checkouts/R/LFQService/inst/run_scripts/lfq_MQ_SampleSizeReport.R ~/Downloads/1330043.zip
```




# How to get started

## 2 grp analysis (TODO)
- See vignettes.

## Parallel design (TODO)
What is parallel design.
- See vignettes.

## Repeated measurements (TODO)
What is a paired measurement.
- See vignettes.

## Contrasts (TODO)
- See vignettes.

# Related resources

- [Triqler](https://github.com/statisticalbiotechnology/triqler)
- [MSQRob](https://github.com/statOmics/MSqRob)
- [PECA/ROPECA](http://bioconductor.org/packages/release/bioc/html/PECA.html)
- PEPA test - analysis including shared peptides
  - [DAPAR](https://github.com/samWieczorek/DAPAR/)
  - [DAPARData](https://github.com/samWieczorek/DAPARdata/)
  - [PEPA validation](https://github.com/ThomasBurger/pepa-validation)

#  Relevant background information

- [Contrasts in R by Rose Maier](https://rstudio-pubs-static.s3.amazonaws.com/65059_586f394d8eb84f84b1baaf56ffb6b47f.html)
- [R Companion](https://rcompanion.org/rcompanion/h_01.html)
- [Extending the Linear Model with R](http://www.maths.bath.ac.uk/~jjf23/ELM/)
- [Bayesian Data Analysis](http://www.stat.columbia.edu/~gelman/book/)
- [Bayesian essentials with R - R package](https://cran.r-project.org/web/packages/bayess/index.html)
- [Contrasts in R - an example vignette by Rose Maier](https://rstudio-pubs-static.s3.amazonaws.com/65059_586f394d8eb84f84b1baaf56ffb6b47f.html)

