---
title: "Limma Backend for prolfqua"
author: "Witold E. Wolski"
date: "`r Sys.Date()`"
output: html_document
vignette: >
  %\VignetteIndexEntry{Limma Backend for prolfqua}
  %\VignetteEncoding{UTF-8}
  %\VignetteEngine{knitr::rmarkdown}
editor_options:
  chunk_output_type: console
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE, message = FALSE, warning = FALSE)
```

# Purpose

This vignette demonstrates the **limma backend** for prolfqua's modelling pipeline. The limma backend is a drop-in alternative to the per-protein `lm` + `squeezeVar` pipeline. Both approaches fit linear models and perform empirical Bayes variance shrinkage, but:

- **prolfqua default** (`strategy_lm` + `Contrasts` + `ContrastsModerated`): fits one `lm()` per protein, then applies prolfqua's own `squeezeVarRob` moderation.
- **limma backend** (`strategy_limma` + `ContrastsLimma`): uses limma's matrix-based `lmFit` + `contrasts.fit` + `eBayes` pipeline directly.

The user-facing API is identical: same formula specification, same contrast specification, same downstream methods (`get_contrasts`, `get_Plotter`, `to_wide`, `merge_contrasts_results`).

# Data preparation

We use simulated protein-level data with 3 groups (A, B, Ctrl) and some missing values.

```{r data}
library(prolfqua)
library(dplyr)

istar <- sim_lfq_data_protein_config(Nprot = 100, weight_missing = 0.3)
lfqdata <- LFQData$new(istar$data, istar$config)
lfqdata$remove_small_intensities()

lt <- lfqdata$get_Transformer()
transformed <- lt$log2()$robscale()$lfq
transformed$rename_response("transformedIntensity")
transformed$factors()
```

# Define contrasts

The same contrast specification is used for both backends.

```{r contrasts}
contr_spec <- c(
  "AvsCtrl" = "group_A - group_Ctrl",
  "BvsCtrl" = "group_B - group_Ctrl",
  "AvsB"    = "group_A - group_B"
)
```

# prolfqua default pipeline

```{r prolfqua_pipeline}
mod_lm <- build_model(transformed, strategy_lm("transformedIntensity ~ group_"))
contr_lm <- prolfqua::Contrasts$new(mod_lm, contr_spec)
contr_moderated <- ContrastsModerated$new(contr_lm)
res_moderated <- contr_moderated$get_contrasts()
```

# Limma backend pipeline

The only difference is using `strategy_limma` + `build_model_limma` + `ContrastsLimma`.

```{r limma_pipeline}
strat_limma <- strategy_limma("transformedIntensity ~ group_")
mod_limma <- build_model_limma(transformed, strat_limma)
contr_limma <- ContrastsLimma$new(mod_limma, contr_spec)
res_limma <- contr_limma$get_contrasts()
```

## Model-level inspection

The `ModelLimma` object supports the same inspection methods as `Model`.

```{r model_inspection}
mod_limma$get_anova() |> head()
mod_limma$get_coefficients() |> head()
mod_limma$anova_histogram()
```

# Comparing results

## Fold changes are identical

Both backends estimate the same linear model, so fold changes (log2 FC) are identical.

```{r compare_fc}
merged <- inner_join(
  select(res_limma, protein_Id, contrast, diff_limma = diff),
  select(res_moderated, protein_Id, contrast, diff_moderated = diff),
  by = c("protein_Id", "contrast")
)

cor(merged$diff_limma, merged$diff_moderated, use = "complete.obs")
```

```{r plot_fc, fig.cap="Fold changes: limma vs. prolfqua moderated. Points lie on the diagonal."}
plot(merged$diff_limma, merged$diff_moderated,
     xlab = "limma logFC", ylab = "prolfqua moderated logFC",
     pch = 20, col = rgb(0, 0, 0, 0.3))
abline(0, 1, col = "red")
```

## P-values differ (different shrinkage)

The p-values differ because limma and prolfqua use different eBayes implementations, but they are highly correlated.

```{r compare_pval}
merged_p <- inner_join(
  select(res_limma, protein_Id, contrast, p_limma = p.value),
  select(res_moderated, protein_Id, contrast, p_moderated = p.value),
  by = c("protein_Id", "contrast")
)

cor(-log10(merged_p$p_limma), -log10(merged_p$p_moderated), use = "complete.obs")
```

```{r plot_pval, fig.cap="P-values: limma vs. prolfqua moderated. High correlation but not identical."}
plot(-log10(merged_p$p_limma), -log10(merged_p$p_moderated),
     xlab = "-log10(p) limma", ylab = "-log10(p) prolfqua moderated",
     pch = 20, col = rgb(0, 0, 0, 0.3))
abline(0, 1, col = "red")
```

# Volcano plots

```{r volcano, fig.cap="Volcano plots from both backends."}
pl_limma <- contr_limma$get_Plotter()
pl_moderated <- contr_moderated$get_Plotter()

gridExtra::grid.arrange(
  pl_limma$volcano()$FDR + ggplot2::ggtitle("limma"),
  pl_moderated$volcano()$FDR + ggplot2::ggtitle("prolfqua moderated"),
  ncol = 1
)
```

# Merging with missing value imputation

`ContrastsLimma` integrates seamlessly with `ContrastsMissing` and `merge_contrasts_results`.

```{r merge_missing}
mC <- ContrastsMissing$new(lfqdata = transformed, contrasts = contr_spec)
merged_result <- merge_contrasts_results(prefer = contr_limma, add = mC)

plotter <- merged_result$merged$get_Plotter()
plotter$volcano()$FDR
```

# Wide format export

```{r wide}
wide <- contr_limma$to_wide()
head(wide)
```

# Two-factor design

The limma backend handles multi-factor designs in the same way.

```{r two_factor}
dd <- sim_lfq_data_2Factor_config(Nprot = 100, with_missing = FALSE)
lProt2 <- LFQData$new(dd$data, dd$config)
lProt2$rename_response("transformedIntensity")

strat2 <- strategy_limma("transformedIntensity ~ Treatment + Background")
mod2 <- build_model_limma(lProt2, strat2)

Contr2 <- c(
  "A_vs_B" = "TreatmentA - TreatmentB",
  "X_vs_Z" = "BackgroundX - BackgroundZ"
)
contr2 <- ContrastsLimma$new(mod2, Contr2)

pl2 <- contr2$get_Plotter()
pl2$volcano()$FDR
```

# strategy_limma options

The `strategy_limma` function accepts `trend` and `robust` arguments that are passed to `limma::eBayes`:

- `trend = TRUE`: allows the prior variance to depend on average intensity (useful when variance is intensity-dependent)
- `robust = TRUE`: uses a robust empirical Bayes procedure (resistant to outlier variances)

```{r options}
strat_robust <- strategy_limma(
  "transformedIntensity ~ group_",
  trend = TRUE,
  robust = TRUE
)
mod_robust <- build_model_limma(transformed, strat_robust)
contr_robust <- ContrastsLimma$new(mod_robust, contr_spec)
head(contr_robust$get_contrasts())
```

# Session Info

```{r}
sessionInfo()
```
