---
title: "Limma Backend for prolfqua"
author: "Witold E. Wolski"
date: "`r Sys.Date()`"
output: html_document
vignette: >
  %\VignetteIndexEntry{Limma Backend for prolfqua}
  %\VignetteEncoding{UTF-8}
  %\VignetteEngine{knitr::rmarkdown}
editor_options:
  chunk_output_type: console
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE, message = FALSE, warning = FALSE)
```

# Purpose

This vignette demonstrates the **limma backend** for prolfqua's modelling pipeline. The limma backend is a drop-in alternative to the per-protein `lm` + `squeezeVar` pipeline. Both approaches fit linear models and perform empirical Bayes variance shrinkage, but:

- **prolfqua default** (`strategy_lm` + `Contrasts` + `ContrastsModerated`): fits one `lm()` per protein, then applies prolfqua's own `squeezeVarRob` moderation.
- **limma backend** (`strategy_limma` + `ContrastsLimma`): uses limma's matrix-based `lmFit` + `contrasts.fit` + `eBayes` pipeline directly.

The user-facing API is identical: same formula specification, same contrast specification, same downstream methods (`get_contrasts`, `get_Plotter`, `to_wide`, `merge_contrasts_results`).

# Data preparation

We use simulated protein-level data with 3 groups (A, B, Ctrl) and some missing values.

```{r data}
library(prolfqua)
library(dplyr)

istar <- sim_lfq_data_protein_config(Nprot = 100, weight_missing = 0.3)
lfqdata <- LFQData$new(istar$data, istar$config)
lfqdata$remove_small_intensities()

lt <- lfqdata$get_Transformer()
transformed <- lt$log2()$robscale()$lfq
transformed$rename_response("transformedIntensity")
transformed$factors()
```

# Define contrasts

The same contrast specification is used for both backends.

```{r contrasts}
contr_spec <- c(
  "AvsCtrl" = "group_A - group_Ctrl",
  "BvsCtrl" = "group_B - group_Ctrl",
  "AvsB"    = "group_A - group_B"
)
```

# prolfqua default pipeline

```{r prolfqua_pipeline}
mod_lm <- build_model(transformed, strategy_lm("transformedIntensity ~ group_"))
contr_lm <- prolfqua::Contrasts$new(mod_lm, contr_spec)
contr_moderated <- ContrastsModerated$new(contr_lm)
res_moderated <- contr_moderated$get_contrasts()
```

# Limma backend pipeline

The only difference is using `strategy_limma` + `build_model_limma` + `ContrastsLimma`.

```{r limma_pipeline}
strat_limma <- strategy_limma("transformedIntensity ~ group_")
mod_limma <- build_model_limma(transformed, strat_limma)
contr_limma <- ContrastsLimma$new(mod_limma, contr_spec)
res_limma <- contr_limma$get_contrasts()
```

## Model-level inspection

The `ModelLimma` object supports the same inspection methods as `Model`.

```{r model_inspection}
mod_limma$get_anova() |> head()
mod_limma$get_coefficients() |> head()
mod_limma$anova_histogram()
```

# Comparing results

## Fold changes are identical

Both backends estimate the same linear model, so fold changes (log2 FC) are identical.

```{r compare_fc}
merged <- inner_join(
  select(res_limma, protein_Id, contrast, diff_limma = diff),
  select(res_moderated, protein_Id, contrast, diff_moderated = diff),
  by = c("protein_Id", "contrast")
)

cor(merged$diff_limma, merged$diff_moderated, use = "complete.obs")
```

```{r plot_fc, fig.cap="Fold changes: limma vs. prolfqua moderated. Points lie on the diagonal."}
plot(merged$diff_limma, merged$diff_moderated,
     xlab = "limma logFC", ylab = "prolfqua moderated logFC",
     pch = 20, col = rgb(0, 0, 0, 0.3))
abline(0, 1, col = "red")
```

## P-values differ (different shrinkage)

The p-values differ because limma and prolfqua use different eBayes implementations, but they are highly correlated.

```{r compare_pval}
merged_p <- inner_join(
  select(res_limma, protein_Id, contrast, p_limma = p.value),
  select(res_moderated, protein_Id, contrast, p_moderated = p.value),
  by = c("protein_Id", "contrast")
)

cor(-log10(merged_p$p_limma), -log10(merged_p$p_moderated), use = "complete.obs")
```

```{r plot_pval, fig.cap="P-values: limma vs. prolfqua moderated. High correlation but not identical."}
plot(-log10(merged_p$p_limma), -log10(merged_p$p_moderated),
     xlab = "-log10(p) limma", ylab = "-log10(p) prolfqua moderated",
     pch = 20, col = rgb(0, 0, 0, 0.3))
abline(0, 1, col = "red")
```

# Volcano plots

```{r volcano, fig.cap="Volcano plots from both backends."}
pl_limma <- contr_limma$get_Plotter()
pl_moderated <- contr_moderated$get_Plotter()

gridExtra::grid.arrange(
  pl_limma$volcano()$FDR + ggplot2::ggtitle("limma"),
  pl_moderated$volcano()$FDR + ggplot2::ggtitle("prolfqua moderated"),
  ncol = 1
)
```

# Merging with missing value imputation

`ContrastsLimma` integrates seamlessly with `ContrastsMissing` and `merge_contrasts_results`.

```{r merge_missing}
mC <- ContrastsMissing$new(lfqdata = transformed, contrasts = contr_spec)
merged_result <- merge_contrasts_results(prefer = contr_limma, add = mC)

plotter <- merged_result$merged$get_Plotter()
plotter$volcano()$FDR
```

# Wide format export

```{r wide}
wide <- contr_limma$to_wide()
head(wide)
```

# Two-factor design

The limma backend handles multi-factor designs in the same way.

```{r two_factor}
dd <- sim_lfq_data_2factor_config(Nprot = 100, with_missing = FALSE)
lProt2 <- LFQData$new(dd$data, dd$config)
lProt2$rename_response("transformedIntensity")

strat2 <- strategy_limma("transformedIntensity ~ Treatment + Background")
mod2 <- build_model_limma(lProt2, strat2)

Contr2 <- c(
  "A_vs_B" = "TreatmentA - TreatmentB",
  "X_vs_Z" = "BackgroundX - BackgroundZ"
)
contr2 <- ContrastsLimma$new(mod2, Contr2)

pl2 <- contr2$get_Plotter()
pl2$volcano()$FDR
```

# strategy_limma options

The `strategy_limma` function accepts `trend` and `robust` arguments that are passed to `limma::eBayes`:

- `trend = TRUE`: allows the prior variance to depend on average intensity (useful when variance is intensity-dependent)
- `robust = TRUE`: uses a robust empirical Bayes procedure (resistant to outlier variances)

```{r options}
strat_robust <- strategy_limma(
  "transformedIntensity ~ group_",
  trend = TRUE,
  robust = TRUE
)
mod_robust <- build_model_limma(transformed, strat_robust)
contr_robust <- ContrastsLimma$new(mod_robust, contr_spec)
head(contr_robust$get_contrasts())
```

# Vooma backend pipeline

The vooma (variance modelling at the observational level) backend extends limma
with observation-level precision weights derived from a mean-variance trend.
This is analogous to voom for RNA-seq but applied to proteomics intensities.
The key idea: instead of assuming equal variance across all observations, vooma
estimates a smooth mean-variance relationship and down-weights observations
with higher expected variance.

The API mirrors the standard limma pipeline — the only difference is calling
`build_model_limma_voom` instead of `build_model_limma`.

```{r vooma_pipeline}
strat_voom <- strategy_limma("transformedIntensity ~ group_")
mod_voom <- build_model_limma_voom(transformed, strat_voom, plot = TRUE)
contr_voom <- ContrastsLimma$new(mod_voom, contr_spec)
res_voom <- contr_voom$get_contrasts()
```

The plot shows `sqrt(sigma)` vs average log-expression. The red curve is the
lowess trend used to derive per-observation weights.

## Comparing limma vs vooma

```{r compare_voom}
merged_voom <- inner_join(
  select(res_limma, protein_Id, contrast, p_limma = p.value, diff_limma = diff),
  select(res_voom, protein_Id, contrast, p_voom = p.value, diff_voom = diff),
  by = c("protein_Id", "contrast")
)

data.frame(
  Metric = c("Fold-change correlation", "P-value correlation"),
  Value = c(
    cor(merged_voom$diff_limma, merged_voom$diff_voom, use = "complete.obs"),
    cor(-log10(merged_voom$p_limma), -log10(merged_voom$p_voom), use = "complete.obs")
  )
) |> knitr::kable(digits = 4)
```

```{r volcano_voom, fig.cap="Volcano plots: limma vs vooma."}
pl_voom <- contr_voom$get_Plotter()
gridExtra::grid.arrange(
  pl_limma$volcano()$FDR + ggplot2::ggtitle("limma"),
  pl_voom$volcano()$FDR + ggplot2::ggtitle("vooma"),
  ncol = 1
)
```


# Limpa backend pipeline

The limpa backend uses limpa's Detection Probability Curve (DPC) for
probabilistic missing value handling, followed by vooma precision weighting
that incorporates the quantification standard errors. The three-step pipeline
is:

1. **`AggregateLimpa`** — estimates the DPC, then quantifies features using
   `limpa::dpcQuant()` (peptide→protein) or `limpa::dpcQuantByRow()`
   (same-level imputation). Produces standard errors and observation counts.
2. **`build_model_limpa`** — fits a vooma model using
   `limpa::voomaLmFitWithImputation()` with the SEs as a bivariate predictor
3. **`ContrastsLimma`** — reused as-is since the output is a standard `MArrayLM`

We show two examples: aggregation to protein level and staying at peptide level.

## Prepare peptide-level data

Both examples start from the same simulated peptide-level dataset.

```{r limpa_data, eval=requireNamespace("limpa", quietly = TRUE)}
istar_pep <- sim_lfq_data_peptide_config(Nprot = 100)
lfq_peptide <- LFQData$new(istar_pep$data, istar_pep$config)
lfq_peptide <- lfq_peptide$get_Transformer()$log2()$lfq

data.frame(
  Property = c("Hierarchy", "Samples", "NAs"),
  Value = c(
    paste(lfq_peptide$hierarchy_keys(), collapse = " > "),
    length(unique(lfq_peptide$data_long()[[lfq_peptide$sample_name()]])),
    sum(is.na(lfq_peptide$data_long()[[lfq_peptide$response()]]))
  )
) |> knitr::kable()
```

## Example 1: Peptide → protein aggregation

### Step 1: Aggregate with AggregateLimpa

`AggregateLimpa` wraps `limpa::dpc()` (DPC estimation) and `limpa::dpcQuant()`
(protein quantification). The output is a protein-level `LFQData` with three
value columns: intensity, standard error (`config$opt_se`), and observation
count (`config$nr_children`). Missing values are probabilistically integrated
out during quantification — no fabricated numbers are imputed.

```{r limpa_agg, eval=requireNamespace("limpa", quietly = TRUE)}
agg <- AggregateLimpa$new(lfq_peptide, "protein")
lfq_protein_limpa <- agg$aggregate()

data.frame(
  Property = c("Input hierarchy", "Output hierarchy", "Response column",
               "SE column", "Nr children column", "NAs in output",
               "DPC beta0", "DPC beta1"),
  Value = c(
    paste(lfq_peptide$hierarchy_keys(), collapse = " > "),
    paste(lfq_protein_limpa$hierarchy_keys(), collapse = " > "),
    lfq_protein_limpa$response(),
    lfq_protein_limpa$get_config()$opt_se,
    lfq_protein_limpa$nr_children_col(),
    sum(is.na(lfq_protein_limpa$data_long()[[lfq_protein_limpa$response()]])),
    round(agg$dpc_result$dpc[1], 3),
    round(agg$dpc_result$dpc[2], 3)
  )
) |> knitr::kable()
```

### Step 2: Build model with build_model_limpa

`build_model_limpa` extracts the SE and observation count matrices from the
aggregated data and passes them to `limpa::voomaLmFitWithImputation()`. The
SEs serve as a second predictor in the vooma variance trend (bivariate: average
intensity + log SE), and the observation counts identify imputed entries for
degree-of-freedom correction.

```{r limpa_prot_model, eval=requireNamespace("limpa", quietly = TRUE)}
response <- lfq_protein_limpa$response()
strat_limpa <- strategy_limpa(paste(response, "~ group_"), plot = TRUE)
mod_limpa_prot <- build_model_limpa(lfq_protein_limpa, strat_limpa)

data.frame(
  Property = c("Model class", "Model name"),
  Value = c(class(mod_limpa_prot)[1], mod_limpa_prot$model_name)
) |> knitr::kable()

mod_limpa_prot$get_coefficients() |> head() |> knitr::kable(digits = 3)
```

### Step 3: Compute contrasts with ContrastsLimma

Since `build_model_limpa` returns a standard `ModelLimma`, we reuse
`ContrastsLimma` directly — same as for the limma and vooma backends.

```{r limpa_prot_contrasts, eval=requireNamespace("limpa", quietly = TRUE)}
contr_limpa_prot <- ContrastsLimma$new(mod_limpa_prot, contr_spec, model_name = "limpa")
res_limpa_prot <- contr_limpa_prot$get_contrasts()
head(res_limpa_prot) |> knitr::kable(digits = 3)
```

### Volcano plot

```{r limpa_prot_volcano, eval=requireNamespace("limpa", quietly = TRUE), fig.cap="Limpa protein-level volcano plot."}
pl_limpa_prot <- contr_limpa_prot$get_Plotter()
pl_limpa_prot$volcano()$FDR
```

### Facade shortcut

The same pipeline is available as a one-liner via `build_contrast_analysis(method = "limpa")`.
The input must be `AggregateLimpa` output (not plain `get_Aggregator()` output),
because the facade needs the SE and observation count columns.

```{r limpa_prot_facade, eval=requireNamespace("limpa", quietly = TRUE)}
fa_limpa_prot <- build_contrast_analysis(
  lfq_protein_limpa, "~ group_", contr_spec, method = "limpa"
)
fa_limpa_prot$get_contrasts() |> head() |> knitr::kable(digits = 3)
```


## Example 2: Peptide-level analysis (no aggregation)

In this example we stay at the peptide level. `AggregateLimpa` with
`impute_only = TRUE` calls `limpa::dpcQuantByRow()`, which fills in missing
values at the peptide level while preserving the full hierarchy. Each peptide
row is treated as its own "protein" (one peptide per group). The output has the
same hierarchy as the input but with no NAs and with SE and observation count
columns attached.

This is useful for peptidoform-level analyses (e.g. PTMs) where aggregation to
protein level is not desired.

### Step 1: Impute at peptide level

```{r limpa_pep_impute, eval=requireNamespace("limpa", quietly = TRUE)}
agg_pep <- AggregateLimpa$new(lfq_peptide, impute_only = TRUE)
lfq_peptide_limpa <- agg_pep$aggregate()

data.frame(
  Property = c("Input hierarchy", "Output hierarchy",
               "NAs in input", "NAs in output", "SE column"),
  Value = c(
    paste(lfq_peptide$hierarchy_keys(), collapse = " > "),
    paste(lfq_peptide_limpa$hierarchy_keys(), collapse = " > "),
    sum(is.na(lfq_peptide$data_long()[[lfq_peptide$response()]])),
    sum(is.na(lfq_peptide_limpa$data_long()[[lfq_peptide_limpa$response()]])),
    lfq_peptide_limpa$get_config()$opt_se
  )
) |> knitr::kable()
```

### Step 2: Build model at peptide level

The peptide-level data still has `protein_Id` and `peptide_Id` in its
hierarchy. We set `hierarchyDepth` so that `hierarchy_keys()` returns only
`protein_Id` — this makes the data "aggregated" from the model's perspective
(one row per protein_Id × peptide_Id × sample, with `subject_id = protein_Id`).

Since each peptide is a separate row in the wide matrix, `build_model_limpa`
fits the model at the peptide level.

```{r limpa_pep_model, eval=requireNamespace("limpa", quietly = TRUE)}
response_pep <- lfq_peptide_limpa$response()
strat_limpa_pep <- strategy_limpa(paste(response_pep, "~ group_"), plot = TRUE)
mod_limpa_pep <- build_model_limpa(lfq_peptide_limpa, strat_limpa_pep)

data.frame(
  Property = c("Model class", "Rows in model (peptides)"),
  Value = c(class(mod_limpa_pep)[1], nrow(mod_limpa_pep$fit$coefficients))
) |> knitr::kable()
```

### Step 3: Compute contrasts

```{r limpa_pep_contrasts, eval=requireNamespace("limpa", quietly = TRUE)}
contr_limpa_pep <- ContrastsLimma$new(mod_limpa_pep, contr_spec, model_name = "limpa_peptide")
res_limpa_pep <- contr_limpa_pep$get_contrasts()

data.frame(
  Property = c("Peptide-level results", "Unique peptides"),
  Value = c(nrow(res_limpa_pep), length(unique(res_limpa_pep$peptide_Id)))
) |> knitr::kable()

head(res_limpa_pep) |> knitr::kable(digits = 3)
```

### Volcano plot

```{r limpa_pep_volcano, eval=requireNamespace("limpa", quietly = TRUE), fig.cap="Limpa peptide-level volcano plot."}
pl_limpa_pep <- contr_limpa_pep$get_Plotter()
pl_limpa_pep$volcano()$FDR
```

### Comparison: protein-level vs peptide-level limpa

```{r limpa_compare, eval=requireNamespace("limpa", quietly = TRUE)}
data.frame(
  Level = c("Protein", "Peptide"),
  Rows = c(nrow(res_limpa_prot), nrow(res_limpa_pep)),
  Proteins = c(length(unique(res_limpa_prot$protein_Id)),
               length(unique(res_limpa_pep$protein_Id))),
  Peptides = c(NA_integer_, length(unique(res_limpa_pep$peptide_Id)))
) |> knitr::kable()
```


# Session Info

```{r session_info}
sessionInfo()
```
