---
title: "Modelling dataset with two Factors"
author: "Witold E. Wolski"
date: "`r Sys.Date()`"
output:
  html_document: default
  pdf_document: default
vignette: |
  %\VignetteIndexEntry{Modelling dataset with two Factors}  
  %\VignetteEncoding{UTF-8}   
  %\VignetteEngine{knitr::rmarkdown}
editor_options: 
  chunk_output_type: console
bibliography:
  - prolfqua.bib
---


```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE, message = FALSE, warning = FALSE)
```

# Purpose
In this vignette we demonstrate how to integrate more than one factor into the linear models. Here, we show how to model the data with two factors plus the interaction thereof. The underlying dataset is generated in a course that is held on a yearly base. The context is that yeast is in one condition grown on glucose and in the other condition yeast is grown on glycerol and ethanol. Here we are looking into the results of two different coursess where of course different people in the wetlab and here even with different LC-MS instruments are involved. It is therefore important to include the batch variable for these two _similar_ datasets. 
For demonstration purposes we are also modelling the interaction between the two explanatory variables _batch_ and _condition_. In this case, having a significant interaction term would mean the protein is expressed more in the Glucose condition in one batch while the same protein is more abundant in the Ethanol condition in the other batch.

# Model Fitting

For more details how the dataset `data_Yeast2Factor` was created we refer you to the prolfquabenchmark vignettes.
Interesting here is the definition of the model. If interaction shall be included in the model a _asterix_ should be used while if no interaction should be taken into account a _plus_ should be used in the model definition.

```{r specifyModel}
conflicted::conflict_prefer("filter", "dplyr")

data_Yeast2Factor <- prolfqua::prolfqua_data("data_Yeast2Factor")
data_Yeast2Factor <- prolfqua::LFQData$new(data_Yeast2Factor$data, prolfqua::old2new(data_Yeast2Factor$config))

pMerged <- data_Yeast2Factor
pMerged$data$Run_ID <- as.numeric(pMerged$data$Run_ID)
pMerged$factors()


formula_Batches <-
  prolfqua::strategy_lm("transformedIntensity ~ condition_ * batch_ ")

# specify model definition
Contrasts <- c("Glucose - Ethanol" = "condition_Glucose - condition_Ethanol",
               "p2370 - p2691" = "batch_p2370 - batch_p2691",
               "Glucose_vs_Ethanol_gv_p2370" = "`condition_Glucose:batch_p2370` - `condition_Ethanol:batch_p2370`",
               "Glucose_vs_Ethanol_gv_p2691" = "`condition_Glucose:batch_p2691` - `condition_Ethanol:batch_p2691`",
               "Interaction" = "`Glucose_vs_Ethanol_gv_p2370` - `Glucose_vs_Ethanol_gv_p2691`")
```


```{r buildModel}
mod <- prolfqua::build_model(pMerged$data, formula_Batches,
  subject_Id = pMerged$config$table$hierarchy_keys() )
```


```{r anovaPvaluePlots, fig.cap="p-value distributions for ANOVA analysis."}
mod$anova_histogram()$plot
```

## ANOVA

Examine proteins with a significant interaction between the two factors treatment and batch.

```{r anovaAnalysis}
ANOVA <- mod$get_anova()
ANOVA |> dplyr::filter(factor == "condition_:batch_") |> dplyr::arrange(FDR.Pr..F.) |> head(5)
protIntSig <- ANOVA |> dplyr::filter(factor == "condition_:batch_") |>
  dplyr::filter(FDR.Pr..F. < 0.05)
protInt <-  pMerged$get_copy()
protInt$data <- protInt$data[protInt$data$protein_Id %in% protIntSig$protein_Id,]
```

These proteins (here the top five) can easily be visualized using the `boxplot` function from the `plotter` objects in prolfqua

```{r fig.with=15, fig.height=15, fig.cap="Proteins with FDR < 0.05 for condition batch interaction in ANOVA."}
ggpubr::ggarrange(plotlist = protInt$get_Plotter()$boxplots()$boxplot)
```

# Compute and analyse with the specified contrasts 

```{r computeModeratedContrasts}
contr <- prolfqua::ContrastsModerated$new(prolfqua::Contrasts$new(mod, Contrasts))
contrdf <- contr$get_contrasts()
```

```{r visualiseResults}
plotter <- contr$get_Plotter()
plotter$volcano()
plotter$ma_plot()
```

## Analyse contrasts with missing data imputation

```{r prepareForGroupAverageImputation}
pMerged$config$table$factor_keys_depth()
contrSimple <- prolfqua::ContrastsSimpleImpute$new(pMerged, Contrasts)
contrdfSimple <- contrSimple$get_contrasts()
pl <- contrSimple$get_Plotter()
pl$histogram_diff()
pl$volcano()
```

## Merge nonimputed and imputed data.

```{r doGroupAverageImputation}

dim(contr$get_contrasts())
dim(contrSimple$get_contrasts())

mergedContrasts <- prolfqua::merge_contrasts_results(prefer = contr, add = contrSimple)$merged
cM <- mergedContrasts$get_Plotter()
plot <- cM$volcano()
plot$FDR
```

### Look at Proteins with significant interaction term.

```{r visualiseSomeProteins, fig.cap="Heatmap for proteins that are quantified only in one condition."}

sigInteraction <- mergedContrasts$contrast_result |> 
  dplyr::filter(contrast == "Interaction" & FDR < 0.2)

protInt <-  pMerged$get_copy()
protInt$data <- protInt$data[protInt$data$protein_Id %in% sigInteraction$protein_Id,]

protInt$get_Plotter()$raster()

hm <- protInt$get_Plotter()$heatmap()
```

```{r interactionHeatmap, fig.cap="Proteinheatmap for proteins with significant Interactions"}
hm
```

The `prolfqua` package is described in [@Wolski2022.06.07.494524].

# Session Info

```{r}
sessionInfo()
```

# References


