---
title: "Modelling with Interactions."
author: "Witold Wolski"
date: "`r format(Sys.time(), '%d %B, %Y')`"
output:
  html_document: default
  pdf_document: default
vignette: |
  %\VignetteIndexEntry{Modelling with Interactions.}  
  %\VignetteEncoding{UTF-8}   
  %\VignetteEngine{knitr::rmarkdown}
editor_options: 
  chunk_output_type: console
---


```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE, message = FALSE, warning = FALSE)
```






```{r}
a <- c(a = 3,b = 4.5, c = 6, d = 7.5, e = 9)
f1 <- list(l1 = 3,l2 = 4.5)
f2 <- list(l1 = 0, l2 = 3) 

a =  f1$l1 + f2$l1
b = f1$l2 + f2$l1
c = f1$l1 + f2$l2
d = f1$l2 + f2$l2
c(a,b,c,d)

```

```{r LoadDataAndConfigure}
datadir <- file.path(find.package("prolfqua") , "samples/maxquant_txt")
inputMQfile <-  file.path(datadir, "tiny2.zip")
inputAnnotation <- file.path(datadir, "annotation_Ionstar2018_PXD003881.xlsx")
startdata <- prolfqua::tidyMQ_ProteinGroups(inputMQfile)
```

Read the sample annotation. The sample annotation must contain the `raw.file` name and the explanatory variables of your experiment, e.g. treatment, timepoint, genetic background, or other factors which you would like to check for confounding.


```{r readAnnotation}
annotation <- readxl::read_xlsx(inputAnnotation)
head(annotation)
annotation <- annotation |> dplyr::filter(sample != "e")
annotation <- annotation |> 
  dplyr::mutate(f1 = dplyr::case_when(sample %in% c("a","c") ~ "l1", TRUE ~ "l2"),
                f2 = dplyr::case_when(sample %in% c("a","b") ~ "l1", TRUE ~ "l2")) |>
  dplyr::arrange(sample)
```

Merge the annotation with quantitative data using `inner_join` joining by `raw.file`.

```{r addAnnotationToData}
startdata <- dplyr::inner_join(annotation, startdata, by = "raw.file")
```

We remove all proteins identified only by a single peptide.


```{r filterForAtLeastTwoPeptides}
startdata <- dplyr::filter(startdata, nr.peptides > 1)
```


Then you need to _tell_ `prolfqua` which columns in the data frame contain what information. You do it using the `AnalysisTableAnnotation` class.

```{r setupConfigs}
atable <- prolfqua::AnalysisTableAnnotation$new()
```

The `AnalysisTableAnnotation` has the following fields that need to be populated:

- fileName
- hierarchy
- factors
- workingIntensity

, and we will discuss in more detail next.


The `fileName` is the column with the raw file names, however for labelled TMT experiments, it can be used to hold the name of the TMT channel.

```{r specifyRawFile}
atable$fileName = "raw.file"
```

The `hierarchy` field describes the structure of the MS data e.g, 

- protein 
- peptides
- modified peptides
- precursor

In case of the MQ proteinGroups file we have data on protein level.

```{r specifyProteinID}
atable$hierarchy[["protein_Id"]] <- c("proteinID")

```

In addition you need to describe the `factors` of the analysis, i.e, the column containing the explanatory variables. 

```{r specifyFactors}
atable$factors[["f1."]] = "f1"
atable$factors[["f2."]] = "f2"

```

We also need to specify the column containing the protein abundances.

```{r specifyIntensity}
atable$setWorkIntensity("mq.protein.intensity")
```

Finally we create the `AnalysisConfiguration` which needs the `AnalysisTableAnnotation` we just created and the `AnalysisParameters`. 

```{r createAnalysisConfig}
config <- prolfqua::AnalysisConfiguration$new(atable)
adata <- prolfqua::setup_analysis(startdata, config)

```

Create the `LFQData` class instance and remove zeros from data (MaxQuant encodes missing values with zero).

```{r removeSmallIntensities}
lfqdata <- prolfqua::LFQData$new(adata, config)
lfqdata$remove_small_intensities()
lfqdata$factors()
```


```{r}
hm <- lfqdata$get_Plotter()$heatmap()
```

```{r plotHeatmap}
hm
```


```{r}
tr <- lfqdata$get_Transformer()
lfqTrans <- tr$log2()$lfq
lfqTrans$get_Plotter()$intensity_distribution_density()
lfqTrans$response()
lfqTrans$rename_response("abundance")

```


# Model Fitting


```{r specifyModel}

formula_Batches <-
  prolfqua::strategy_lm("abundance ~ f1. * f2. ")

# specify model definition
modelName  <- "Model"
DEBUG <- TRUE

Contrasts <- c("f1.l1vsf1.l2" = "f1.l1 - f1.l2",
               "f2.l1vsf2.l2" = "f2.l1 - f2.l2",
               "f1l1vsf1l2_gv_f2.l1" = "`f1.l1:f2.l1` - `f1.l2:f2.l1`",
               "f1l1vsf1l2_gv_f2.l2" = "`f1.l1:f2.l2` - `f1.l2:f2.l2`",
               "Interaction" = "`f1l1vsf1l2_gv_f2.l1` - `f1l1vsf1l2_gv_f2.l2`"
               )

```


```{r buildModel}

mod <- prolfqua::build_model(
  lfqTrans,
  formula_Batches)

```



```{r anovaPvaluePlots, fig.cap="p-value distributions for ANOVA analysis."}
mod$anova_histogram(what = "FDR.Pr..F.")

```

## ANOVA

Examine proteins with a significant interaction between the two factors treatment and batch.

```{r anovaAnalysis}
ANOVA <- mod$get_anova()
ANOVA |> dplyr::filter(factor == "f1.:f2.") |> dplyr::arrange(FDR.Pr..F.) |> head(5)
ANOVA$factor |> unique()
protIntSig <- ANOVA |> dplyr::filter(factor == "f1.") |>
  dplyr::filter(FDR.Pr..F. < 0.25)
protInt <-  lfqTrans$get_copy()
protInt$data <- protInt$data[protInt$data$protein_Id %in% protIntSig$protein_Id,]
```


```{r fig.with=15, fig.height=15, fig.cap="Proteins with FDR < 0.5 for condition batch interaction in ANOVA."}
ggpubr::ggarrange(plotlist = protInt$get_Plotter()$boxplots()$boxplot)
```

# Compute contrasts

```{r computeModeratedContrasts}
contr <- prolfqua::ContrastsModerated$new(prolfqua::Contrasts$new(mod, Contrasts))
#contr$get_contrasts_sides()
contrdf <- contr$get_contrasts()
```

```{r}
plotter <- contr$get_Plotter()
plotter$volcano()
plotter$ma_plot()

```



## Annalyse contrasts with missing data imputation

```{r}
lfqTrans$config$table$factorDepth <- 2
#ContrastsSimpleImpute$debug("get_contrasts")
contrSimple <- prolfqua::ContrastsSimpleImpute$new(lfqdata = lfqTrans, Contrasts)
contrdfSimple <- contrSimple$get_contrasts()
#na.omit(contrdfSimple)
pl <- contrSimple$get_Plotter()
pl$histogram_diff()
pl$volcano()
```


## Merge nonimputed and imputed data.

```{r}

dim(contr$get_contrasts())
dim(contrSimple$get_contrasts())

mergedContrasts <- prolfqua::addContrastResults(prefer = contr, add = contrSimple)$merged
cM <- mergedContrasts$get_Plotter()
plot <- cM$volcano()
plot$FDR
```




