---
title: "QC and Sample Size Estimation"
author: "Witold Wolski"
date: "14/12/2020"
output: html_document
vignette: >
  %\VignetteIndexEntry{QC and Sample Size Estimation} 
  %\VignetteEncoding{UTF-8}
  %\VignetteEngine{knitr::rmarkdown}
editor_options: 
  chunk_output_type: console
---


```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE, message = FALSE, warning = FALSE)
```


# Loading data


```{r createConfig}
library(prolfqua)
library(tidyverse)


```


Load the data from MaxQuant proteinGroups.txt file.

```{r LoadDataAndConfigure}
datadir <- file.path(find.package("prolfquaData") , "quantdata")
inputMQfile <-  file.path(datadir, "MAXQuant_ComboCourse_p2691_March_2018_WU183012.zip")
inputAnnotation <- file.path(datadir, "annotation_ComboCourse_p2691_March_2018_WU183012.xlsx")

startdata <- prolfqua::tidyMQ_ProteinGroups(inputMQfile)
startdata$proteinID <- gsub(";.*$","", startdata$majority.protein.ids)
```

Read the annotation with the explanatory variables.

```{r}

annotation <- readxl::read_xlsx(inputAnnotation)
annotation$experiment = "p2691"
head(annotation)

```


```{r}
startdata <- inner_join(annotation, startdata, by = "raw.file")
head(startdata)
```

For the sample size estimation we will use bio-reps from the _Ethanol_ condition. 
We also remove all proteins identified only by a single peptide.


```{r}
startdata <- startdata %>% dplyr::filter(condition  == "Ethanol")
startdata <- startdata %>% dplyr::filter(nr.peptides > 1)

```


We need to _tell_ `prolfqua` at which columns in the startdata frame contain what information. This is done using the `AnalysisTableAnnotation` class.

The `hierarchy` describes the structure of the MS data. In this case we have only protein level measurements.
In addition you need to describe the factors of the analysis, i.e, the column containing the explanatory variables.
You also need to specify the column containing the protein identifiers and the protein intensities.

```{r}

atable <- AnalysisTableAnnotation$new()
atable$fileName = "raw.file"
atable$hierarchy[["protein_Id"]] <- c("proteinID")
atable$hierarchyDepth <- 1
atable$setWorkIntensity("mq.protein.intensity")
atable$factors[["condition_"]] = "condition"
atable$factors[["batch_"]] = "experiment"
atable$factors[["Run_ID"]] = "Run_ID"
atable$factorDepth <- 1


config <- AnalysisConfiguration$new(atable)


adata <- setup_analysis(startdata, config)

```


Create the LFQData class instance, remove zeros from data.

```{r filterdata}
lfqdata <- LFQData$new(adata, config)
lfqdata$remove_small_intensities()
```

You can convert the data into a data frame in wide format, where the intensities of each sample occupy their own columns.

```{r}
lfqdata$to_wide()$data[1000:1003,1:3]
```


## Visualization of not normalized data

```{r makeMissingHeatmap}
lfqplotter <- lfqdata$get_Plotter()
lfqplotter$intensity_distribution_density()
```


### Visualization of missing data

```{r}
nah <- lfqplotter$NA_heatmap()
```

```{r printMissingHeatmap, fig.cap="Heatmap, black - missing protein intensities, white - present"}
nah
```



```{r fig.cap="# of proteins with 0,1,...N missing values"}
lfqdata$get_Summariser()$plot_missingness_per_condition()
```

```{r fig.cap="Cumulative sum of the # of proteins with 0,1,...N missing values"}
lfqdata$get_Summariser()$plot_missingness_per_condition_cumsum()
```

```{r fig.cap="Intensity distribution of proteins depending on # of missing values"}
lfqplotter$missigness_histogram()
```

# Computing standard deviations, mean and CV.


```{r PlotCVDistributions}
stats <- lfqdata$get_Stats()
prolfqua::table_facade( stats$stats_quantiles()$wide, paste0("quantile of ",stats$stat ))
```


```{r fig.cap="Distribution of CV's"}
stats$density_median()

```


```{r fig.cap="Scatter plot of standard deviation vs mean"}
stdm_raw <- stats$stdv_vs_mean(size = 10000) + scale_x_log10() + scale_y_log10()
stdm_raw
```


## Normalize protein intensities

We normalize the data by first $\log_2$ transforming it and then z-scaling.
The $log_2$ stabilizes the variance, while the z-scaling removes systematic differences from the samples.

```{r normalizedata}
lt <- lfqdata$get_Transformer()
transformed <- lt$log2()$robscale()$lfq
transformed$config$table$is_intensity_transformed

```

```{r plotDensities, fig.cap="Normalized intensities."}
pl <- transformed$get_Plotter()
pl$intensity_distribution_density()
```

```{r plotScatterMatrix, fig.cap = "Scatterplot matrix"}
pl$pairs_smooth()
```

```{r createHeatmap}
p <- pl$heatmap()
```

```{r plotheatmap, fig.cap="Heatmap, Rows - proteins, Columns - samples", fig.align=5, fig.height=5}
p
```

We can also look at the correlation among the samples.

```{r heatmapCor, fig.cap="Heatmap based on sample correlations, Rows - samples, Columns - samples", fig.align=5, fig.height=5}
hc <- pl$heatmap_cor()
```

```{r plotHeatmapCor}
hc
```


```{r showStandardDeviations}
stats <- transformed$get_Stats()
prolfqua::table_facade(stats$stats_quantiles()$wide, "Standard deviations")

```


```{r sddensity, fig.cap="Density of the standard deviation"}
stats$density_median()
```

```{r experiment, include = FALSE}
chsq <- na.omit(stats$statsdf$sd^2)
plot(density(chsq[chsq < 10]))
x <- seq(0,50, length = 1000)
lines(x*median(chsq) , dchisq(x, df = 4 ), col = 2)

```

Check for heteroskedasticity. After transformation the sd should be independent of the mean intensity.

```{r checkForHeteroskedasticity, fig.cap="Scatter plot of sd vs mean of protein intensity"}
stdm_trans <- stats$stdv_vs_mean(size = 10000) + scale_x_log10() + scale_y_log10()
stdm_trans
#gridExtra::grid.arrange(stdm_raw, stdm_trans, nrow=1)
```



## Estimate sample size


```{r fig.cap="Empirical cumulative density function of the standard deviation."}
stats$density(ggstat = "ecdf")
  
```



```{r estimateSampleSizes}
sampleSize <- stats$power_t_test_quantiles() %>% 
  dplyr::filter(condition_ != "All")
prolfqua::table_facade(sampleSize, "Sample sizes. delta - Effect size, N - samplesize")
```

```{r fig.cap="Estimated sample sizes for various FC levels and various quantiles of the standard deviation."}
sampleSize %>%
  ggplot(aes(x = factor(probs) , y = N)) +
  facet_wrap(~delta) +
  geom_bar(stat = "identity")

```

Getting statistics for each protein.

```{r stats, eval = TRUE}
head(stats$stats())

```


Getting the sample size for each protein.

```{r statsWithN, eval = TRUE}

x <- stats$power_t_test()
x <- x %>% dplyr::filter(condition_ == "Ethanol") %>% arrange(desc(N),protein_Id)
prolfqua::table_facade(x[1000:1007,], caption = "Sample size for each protein")

```


```{r fig.cap="Distribution of the required sample sizes for three fold change thresholds for all the proteins."}
x %>% ggplot(aes(x = N)) + geom_histogram() + facet_wrap(~delta) + xlim(0,100)

```




