---
title: "Analyse correlation among precursors"
author: "Functional Genomics Center Zurich"
date: "`r format(Sys.time(), '%d %B, %Y')`"
output:
  pdf_document: default
  html_document: default
vignette: >
  %\VignetteIndexEntry{Analyse correlations among precursors}
  %\VignetteEncoding{UTF-8}
  %\VignetteEngine{knitr::rmarkdown}
editor_options: 
  chunk_output_type: console
---


This vignette describes how QValue filtering, missing value filtering and than `correlation filtering` can be implemented.




```{r setup, include=FALSE}
knitr::opts_chunk$set(
  echo = TRUE,
  message = FALSE,
  warning = FALSE,
  fig.width = 10,
  fig.height = 10
)

```


First we load the required packages.

```{r}
rm(list = ls())

library(conflicted)
library(tidyverse)
library(dplyr)
library(ggplot2)
library(LFQService)
conflict_prefer("filter", "dplyr")

```


Next the configuration for Spectronaut is clreated and factor variables are defined.

```{r}
config <- create_config_Spectronaut_Peptide()
config$table$factors[["coding"]] = "coding"
config$table$factors[["sex"]] = "sex"
config$table$factors[["age"]] = "age"
config$table$factors[["Sample_id"]] = "Sample.Name"

x <- R6extractValues(config)

```


```{r include=FALSE, eval=FALSE, echo=FALSE}

specnautconfig <- function(){
spectronautDIAData250_config <- LFQService::create_config_Spectronaut_Peptide(
isotopeLabel="Isotope.Label",
 ident_qValue="EG.Qvalue")
spectronautDIAData250_config$table$factors[["coding"]] = "coding"
spectronautDIAData250_config$table$factors[["sex"]] = "sex"
spectronautDIAData250_config$table$factors[["age"]] = "age"
spectronautDIAData250_config$table$factors[["Sample_id"]] = "Sample.Name"
return(spectronautDIAData250_config)
}
config <- specnautconfig()

config$table$factors
config$table$hierarchy
config$table$valueVars()
config$table$idVars()

#'
data <- LFQService::spectronautDIAData250_A
colnames(data$data)
data$data$EG.TotalQuantity..Settings.
pIDS <- sample(unique(data$data$PG.ProteinAccessions),50)


colnames(data$data)
toremove <- c("EG.TotalQuantity..Settings.",
 "EG.UsedForPeptideQuantity","EG.UsedForProteinGroupQuantity")
dataval <- data$data %>% select(-all_of(toremove))
dataval <- dataval %>% filter(PG.ProteinAccessions %in% pIDS)


setup <- function(data, config){
 res <- setup_analysis(data,config)
 res
}
#'
spectronautDIAData250_A <- list(data = dataval, config_f = specnautconfig, analysis=setup )
spectronautDIAData250_A$analysis(spectronautDIAData250_A$data, spectronautDIAData250_A$config_f())
#usethis::use_data( spectronautDIAData250_A , overwrite = TRUE )

```


Given a sample DIA dataset and the configuration the `setup_analysis` function assures that all required columns are present.

```{r}
longFormat <- LFQService::spectronautDIAData250_A$data
"PEP.StrippedSequence" %in% colnames(longFormat)
longFormat$Isotope.Label <- "Light"

config_tmp <- config$clone(deep = TRUE)
longNoDecoy <- setup_analysis(longFormat, config)

```


# QValue Filtering

QValue filtering thresholds are specified in the configuration.
First we remove precursors with QValue larger than `qVal_individual_threshold`.

```{r}
config$parameter$qVal_individual_threshold
longNoDecoy <- removeLarge_Q_Values(longNoDecoy, config)
longQSummaries <- summariseQValues(longNoDecoy, config)

knitr::kable(head(LFQService::interaction_missing_stats(longQSummaries, config)$data))

```


```{r}
knitr::kable(LFQService::hierarchy_counts(longQSummaries, config))
```

Then we remove precrusors that do not pass the  `qVal_experiment_threshold`, that is we require that in at least one sample a precursor was observed with a QValue lower than that threshold.

```{r}
config$parameter$qVal_experiment_threshold 
qvalFilt <- longQSummaries %>% 
  filter_at( "srm_QValueMin" , all_vars(. < config$parameter$qVal_experiment_threshold )   )
LFQService::hierarchy_counts(qvalFilt, config)
```

# NA filtering

The next code snipped computes the number of observed values for each precursor, than keeps only the precursors with more than $20$ observations.

```{r}
longQNASummaries <- rankPrecursorsByNAs(qvalFilt, config)
longQNASummaries <- longQNASummaries %>%
  dplyr::filter(srm_NrNotNAs > 20)
knitr::kable(LFQService::hierarchy_counts(longQNASummaries, config))

```


Finally we plot a heatmap of the cleaned up data.

```{r}
res <- plot_heatmap_cor(longQNASummaries, config)
print(res)
```


# Remove single hit wonders


```{r}

qvalFiltV <- LFQService::filter_proteins_by_peptide_count(longQNASummaries,config)
qvalFiltV
```


# Correlation Filtering

First mark de-correlated transitions using `markDecorrelated`.


```{r markdecorellated}
#config$table$popWorkIntensity()
qvalFiltV <- transform_work_intensity(qvalFiltV$data, config, log2)
xx <- markDecorrelated(qvalFiltV, config, minCorrelation = 0.5 )
mean(xx$srm_decorelated)

```

```{r shownotcorrelated}
nested <- xx %>% group_by_at(config$table$hkeysDepth()) %>% tidyr::nest()

plot_hierarchies_line(nested$data[[1]], nested$protein_Id[[1]], config) + 
  geom_line(aes_string(linetype = "srm_decorelated"), lwd=0.2) 
plot_hierarchies_line(nested$data[[2]], nested$protein_Id[[2]], config) + 
  geom_line(aes_string(linetype = "srm_decorelated"))
plot_hierarchies_line(nested$data[[3]], nested$protein_Id[[3]], config) + 
  geom_line(aes_string(linetype = "srm_decorelated"))
```

Now remove the de-correlated transitions.

```{r summarizeDataAfterCorrelationRemoval}
knitr::kable(hierarchy_counts(xx, config))
qvalFiltCorr <- dplyr::filter(xx, srm_decorelated == FALSE)
knitr::kable(hierarchy_counts(qvalFiltCorr, config))
```

```{r clusterOfCorrelatedData}
print(plot_heatmap_cor(qvalFiltCorr, config))
```

### Rank precursors by intensity and aggregate top 3 precursors 

```{r rankprec}
qvalFiltImputed <- impute_correlationBased(qvalFiltCorr, config)
qvalFiltImputed <- rankPrecursorsByIntensity(qvalFiltImputed, config)

```



```{r aggregate}
mean_na <- function(x, name = FALSE){if(name){return("mean_na")}; mean(x,na.rm = TRUE)}
proteinIntensities <- aggregateTopNIntensities(qvalFiltImputed,
                                               config,
                                               .func = mean_na,
                                               N = 3)$data

```


```{r doplottingOFProtein}
# nest transtions
xnested <- qvalFiltImputed %>% group_by_at(config$table$hkeysDepth()) %>% tidyr::nest()


plot_hierarchies_line(xnested$data[[1]], xnested$protein_Id[1],config)
plot_hierarchies_line(xnested$data[[3]], xnested$protein_Id[3],config)
#plot_hierarchies_line(xnested$data[[4]], xnested$protein_Id[4],config)
# nest protein quants.
figs3 <- xnested %>% dplyr::mutate(plotlog = map2(data, UQ(sym(config$table$hierarchyKeys()[1])),plot_hierarchies_line, config))
figs3

ff <- proteinIntensities %>% 
  group_by_at(config$table$hierarchyKeys()[1]) %>% 
  tidyr::nest(.key = "topNIntensity")

figs4 <- dplyr::inner_join(figs3,ff, by = config$table$hierarchyKeys()[1])

plot_hierarchies_add_quantline(figs4$plotlog[[1]], figs4$topNIntensity[[1]], "srm_mean_na_3", config)
plot_hierarchies_add_quantline(figs4$plotlog[[2]], figs4$topNIntensity[[2]], "srm_mean_na_3", config)
plot_hierarchies_add_quantline(figs4$plotlog[[3]], figs4$topNIntensity[[3]], "srm_mean_na_3", config)


nrPep <- qvalFiltImputed %>% 
  select_at(c(config$table$hierarchyKeys()[1], "nr_peptide_Id_IN_protein_Id")) %>% 
  distinct()

nrPep

```


