---
title: "Simulate Peptide level Data"
author: "Witold E. Wolski"
date: "`r Sys.Date()`"
output: html_document
vignette: >
  %\VignetteIndexEntry{Simulate Peptide level Data} 
  %\VignetteEncoding{UTF-8}
  %\VignetteEngine{knitr::rmarkdown}

editor_options: 
  chunk_output_type: console
bibliography:
  - prolfqua.bib
---


## Dimulate data

```{r setup, include=FALSE}
knitr::opts_chunk$set(warning = FALSE, message = FALSE) 

```


For proteins:
- the proteins have a FC either equal 1, 0. or -1, 10% have 1 80% have 0 and 10% have -1.

What we however are measuring are peptide spectrum matches. Let's assume we observing peptides.

For peptides:

- The transformed protein abundances have a log normal distribution with `meanlog = log(20)`, and `sd = log(1.2)`.
- The number of peptides per protein follow a geometric distribution, $N_{pep} \sim Geo(p)$ with $p = 0.3$
- The peptide abundances of a protein have log normal distribution with `meanlog = log(proteinabundance)` and `sd = log(1.2)`
- The log2 intensities of a peptide within a group follow a normal distribution distribution  $I_{pep} \sim LogNormal(\mu,\sigma) $, where $\mu$ is the peptide abundance and $\sigma$


```{r simulate_peptide_abundances}
peptideAbundances <- prolfqua::sim_lfq_data(PEPTIDE = TRUE)
```


# Analyse simulated data with prolfqua

```{r build_lfqdata}
library(prolfqua)

config <- AnalysisConfiguration$new()
config$file_name = "sample"
config$factors["group_"] = "group"
config$hierarchy[["protein_Id"]] = "proteinID"
config$hierarchy[["peptide_Id"]] = "peptideID"
config$set_response("abundance")

adata <- setup_analysis(peptideAbundances, config)

lfqdata <- prolfqua::LFQData$new(adata, config)
lfqdata$is_transformed(TRUE)

lfqdata$remove_small_intensities(threshold = 1)
lfqdata$filter_proteins_by_peptide_count()

lfqdata$factors()
```

```{r plot_heatmap}
pl <- lfqdata$get_Plotter()
lfqdata$hierarchy_counts()
lfqdata$relevant_hierarchy_keys()

pl$heatmap()
```



```{r plot_intensity_distribution}
pl$intensity_distribution_density()
```


# Fit peptide model

```{r fit_peptide_model}
formula_Condition <-  strategy_lm("abundance ~ group_")
lfqdata$set_config_value("hierarchy_depth", 2)

# specify model definition
modelName  <- "Model"
contr_spec <- c("B_over_Ctrl" = "group_B - group_Ctrl",
               "A_over_Ctrl" = "group_A - group_Ctrl")
lfqdata$subject_id()

mod <- prolfqua::build_model(
  lfqdata,
  formula_Condition)
aovtable <- mod$get_anova()
mod$anova_histogram()$plot
```

```{r significant_proteins}
xx <- aovtable |> dplyr::filter(FDR < 0.05)
signif <- lfqdata$get_copy()
signif$set_data(signif$data_long() |> dplyr::filter(protein_Id %in% xx$protein_Id))
signif$get_Plotter()$heatmap()
```

# Aggregate data

```{r aggregate_data}
lfqdata$set_config_value("hierarchy_depth", 1)
protData <- lfqdata$get_Aggregator()$aggregate()
```

```{r protein_model_contrasts}
protData$response()
formula_Condition <-  strategy_lm("medpolish ~ group_")

mod <- prolfqua::build_model(
  protData,
  formula_Condition)

contr <- prolfqua::Contrasts$new(mod, contr_spec)
v1 <- contr$get_Plotter()$volcano()
v1$FDR

ctr <- contr$get_contrasts()

```

# Session Info

```{r session_info}
sessionInfo()
```






