---
title: "Proprocessing the Ionstar dataset"
author: "FGCZ - (Draft)"
date: "`r format(Sys.time(), '%d %B, %Y')`"
output: pdf_document
papersize: a4
geometry: margin=.5in
vignette: >
  %\VignetteIndexEntry{Proprocessing the Ionstar dataset} 
  %\VignetteEncoding{UTF-8}
  %\VignetteEngine{knitr::rmarkdown}
  
---

# Example of subset normalization of the ionstar dataset.


Please download and install the `prolfquaData` package from github

```{r setup, include=FALSE}
knitr::opts_chunk$set(
echo = TRUE,
message = FALSE,
warning = FALSE,
fig.width = 10,
fig.height = 10
)
```


```{r}
rm(list = ls())
library(conflicted)
library(prolfqua)
library(tidyverse)
library(dplyr)
conflict_prefer("filter", "dplyr")

```


We start by loading the IonStar dataset and the annotation from the `prolfquaData` package. The method `add_annotation` adds the annotation to the data.

```{r}
datadir <- file.path(find.package("prolfquaData") , "quantdata")
inputMQfile <-  file.path(datadir, "MaxQuant_Ionstar2018_PXD003881.zip")
inputAnnotation <- file.path(datadir, "annotation_Ionstar2018_PXD003881.xlsx")

mqdata <- tidyMQ_Peptides_Config(inputMQfile)
names
annotation <- readxl::read_xlsx(inputAnnotation)


res <- add_annotation(
  mqdata$data,
  annotation,
  fileName = "raw.file"
)

```



```{r}

mqdata$config$table$factors[["dilution."]] = "sample"
mqdata$config$table$factors[["run_Id"]] = "run_ID"
mqdata$config$table$factorDepth <- 1

mqdata$data <- setup_analysis(res, mqdata$config)

```



```{r}
datafilt <- prolfqua::filter_proteins_by_peptide_count(mqdata$data, mqdata$config)$data
```

Create a sample of N proteins to speed up computations.

```{r}
N <- 200
set.seed(2020)
hk <- mqdata$config$table$hkeysDepth()
protein_Ids <- mqdata$data %>%
  select(all_of(hk)) %>%
  distinct() %>%
  sample_n(size = N)

```


## Normalize data using HUMAN proteins only



```{r filterForHumans}

lfqdata <- LFQData$new(datafilt, mqdata$config)
tr <- lfqdata$get_Transformer()

subset_h <- lfqdata$get_copy()
subset_h$data <- subset_h$data %>% filter(grepl("HUMAN", protein_Id))
lfqdataNormalized <- tr$log2_robscale_subset(lfqsubset = subset_h)

```


The figures below show the intensity distribution before and after normalization.

```{r PlotIntensityDistributions}
before <- lfqdata$get_Plotter()
before$intensity_distribution_density()

after <- lfqdataNormalized$get_Plotter()
after$intensity_distribution_density()

```


# Create result object for peptides

We collect all the relevant data for the Ionstar data in an R6 object.

```{r R6Ionstar, echo =  FALSE}
IonstarData <- R6::R6Class(
  "IonstarData",
  public = list(
    data = NULL,
    config = NULL,
    data_N = NULL,
    config_N = NULL,
    
    initialize = function(data, config, data_N, config_N){
      self$data = data
      self$config = config
      self$data_N = data_N
      self$config_N = config_N
    },
    Pep = function(){
      return(list(data = self$data, config = self$config$clone(deep = TRUE)))
    },
    filtered = function(){
      data <- prolfqua:::filter_proteins_by_peptide_count( self$data ,  self$config )$data
      return(list(data = data, config = self$config$clone(deep = TRUE)))
    },
    normalized = function(){
      return(list(data = self$data_N, config = self$config_N))
    },
    subset_normalized = function(){
      return(list(data = self$data_N, config = self$config_N))
    }
  )
)
lfqNormSubset <- lfqdataNormalized$get_copy()

lfqNormSubset$data <- inner_join(protein_Ids, lfqNormSubset$data)
mqdataSubset <- mqdata
mqdataSubset$data  <- inner_join(protein_Ids, mqdata$data)

ionstar <- IonstarData$new(mqdataSubset$data, mqdataSubset$config, 
                           lfqNormSubset$data, lfqNormSubset$config)

#usethis::use_data(ionstar, overwrite = TRUE)

```


# Protein intensities using Tukeys Median Polish

We will be using the `LFQDataAggregator` class. To estimate protein intensities using Tukey's median polish we need to use normalized data.
The figure below shows the the protein estimates (dashed line) and the peptide intensities.

```{r}
lfqAggMedpol <- LFQDataAggregator$new(lfqdataNormalized, "protein")
lfqAggMedpol$medpolish()
xx <- lfqAggMedpol$plot()
xx$plots[[1]]
```



```{r echo=FALSE}
dataIonstarProtein_subsetNorm <- list(
  data = inner_join(protein_Ids,lfqAggMedpol$lfq_agg$data),
  config = lfqAggMedpol$lfq_agg$config
)

#usethis::use_data(dataIonstarProtein_subsetNorm, overwrite = TRUE)

```

We can also estimate the protein intensities using the top N most abundant peptides. In this case we are using the untransformed protein intensities. The figure below shows the estimated protein intensities.

```{r}
lfqAggregator <- LFQDataAggregator$new(lfqdata, "protein_topN")
lfqAggregator$mean_topN()
topN <- lfqAggregator$plot()
topN$plots[[1]]

```


# Fit model to protein level data

```{r}
DEBUG <- FALSE

Contrasts <- c(
  "dilution_(9/3)_3" =   "dilution.e - dilution.a",
  "dilution_(9/4.5)_2" =   "dilution.e - dilution.b",
  "dilution_(9/6)_1.5" =   "dilution.e - dilution.c",
  "dilution_(9/7.5)_1.2" =   "dilution.e - dilution.d",
  
  "dilution_(7.5/3)_2.5" =   "dilution.d - dilution.a",
  "dilution_(7.5/4.5)_1.6(6)" =   "dilution.d - dilution.b",
  "dilution_(7.5/6)_1.25" =   "dilution.d - dilution.c",
  
  "dilution_(6/3)_2" =   "dilution.c - dilution.a",
  "dilution_(6/4.5)_1.3(3)" =   "dilution.c - dilution.b",
  
  "dilution_(4.5/3)_1.5" =   "dilution.b - dilution.a"
)


tt <- Reduce(rbind, strsplit(names(Contrasts),split = "_"))
tt <- data.frame(tt)[,2:3]
colnames(tt) <- c("ratio" , "expected fold-change")
tt <- tibble::add_column(tt, contrast =  Contrasts, .before = 1)
prolfqua::table_facade(tt, caption = "All possible Contrasts given 5 E. coli dilutions of the Ionstar Dataset")

```


```{r usedContrasts}
relevantContrasts <- c("dilution_(9/7.5)_1.2",
                       "dilution_(7.5/6)_1.25",
                       "dilution_(6/4.5)_1.3(3)",
                       "dilution_(4.5/3)_1.5" )

tt <- Reduce(rbind, strsplit(relevantContrasts,split = "_"))
tt <- data.frame(tt)[,2:3]
colnames(tt) <- c("ratio" , "expected fold-change")
tt <- tibble::add_column(tt, contrast =  Contrasts[names(Contrasts) %in% relevantContrasts], .before = 1)
prolfqua::table_facade(tt, caption = "Contrasts used for benchmark.")
relevantContrasts <- Contrasts[names(Contrasts) %in% relevantContrasts]

```


# Fitting all the proteins based on protein intensities with linear models. 

```{r buildModelLM}
protLFQ <- lfqAggMedpol$lfq_agg

lmmodel <- "~ dilution."
lmmodel <- paste0(protLFQ$config$table$getWorkIntensity() , lmmodel)

modelFunction <- make_custom_model_lm( lmmodel, model_name = "Model")

mod <- prolfqua::build_model(protLFQ$data, modelFunction)
mod$coef_pairs()$plot

```

```{r computeContrastsModerated}
contr <- prolfqua::ContrastsModerated$new(mod, relevantContrasts)
contrdf <- contr$get_contrasts()
cp <- contr$get_Plotter()
cp$volcano()
```

# Benchmarking


```{r benchmark}
ttd <- prolfqua::ionstar_bench_preprocess(contrdf)
medpol_benchmark <- make_benchmark(ttd$data,
                                   model_description = "med. polish and lm. density",
                                   model_name = "prot_med_lm",
                                   FDRvsFDP = list(list(sc = "FDR", desc = FALSE),
                                                  list(sc = "FDR.moderated", desc = FALSE))
)
medpol_benchmark$data()
medpol_benchmark$plot_score_distribution()
medpol_benchmark$plot_ROC()
medpol_benchmark$plot_FDRvsFDP()

```

# Imputation

```{r benchmarkImputation, eval=TRUE}
contrsimple <- prolfqua::ContrastsSimpleImpute$new(lfqdataNormalized, relevantContrasts)

cs <- contrsimple$get_contrasts()
cs <- cs %>% rename(estimate = .data$estimate_median)
ttd <- prolfqua::ionstar_bench_preprocess(cs)
ttd$data
medpol_benchmark <- make_benchmark(ttd$data,
                                   model_description = "imputed data",
                                   model_name = "pep_imputed_nonparam",
                                   toscale = NULL,
                                   benchmark = list(
                                     list(sc = "estimate", desc = TRUE),
                                     list(sc = "statistic", desc = TRUE)
                                   ),
                                   FDRvsFDP = NULL)

medpol_benchmark$data()
medpol_benchmark$plot_score_distribution()
medpol_benchmark$plot_ROC()
x <- medpol_benchmark$pAUC_summaries()
x$barp

```

