---
title: "Data Preprocessing of Yeast data from the Combined Course FGCZ"
author: "Witold Wolski"
date: "06/10/2020"
output: html_document
vignette: >
  %\VignetteIndexEntry{Data Preprocessing of Yeast data from the Combined Course FGCZ} 
  %\VignetteEncoding{UTF-8}
  %\VignetteEngine{knitr::rmarkdown}

---

```{r setup, include=FALSE}
knitr::opts_chunk$set(
echo = TRUE,
message = FALSE,
warning = FALSE,
fig.width = 10,
fig.height = 10
)
```



Create MaxQuant Configuration (only works with `LFQService::tidyMQ_Peptides` function)

```{r}
library(LFQService)
library(tidyverse)
conflicted::conflict_prefer("filter", "dplyr")


# creates default configuration
create_MQ_peptide_Configuration <- function(ident_qValue = "pep",
                                            intensity = "peptide.intensity"
){
  atable <- AnalysisTableAnnotation$new()
  atable$fileName = "raw.file"
  # measurement levels.
  atable$hierarchy[["protein_Id"]] <- c("top_protein")
  atable$hierarchy[["peptide_Id"]] <- c("sequence")
  atable$hierarchyDepth <- 1
  #
  atable$ident_qValue = ident_qValue
  atable$setWorkIntensity(intensity)
  
  anaparam <- AnalysisParameters$new()
  anaparam$min_peptides_protein <- 2
  configuration <- AnalysisConfiguration$new(atable, anaparam)
  
  return(configuration)
}
```

# Prepare p2691 dataset

```{r}
datadir <- file.path(find.package("prolfquaData") , "quantdata")
inputMQfile <-  file.path(datadir, "MAXQuant_ComboCourse_p2691_March_2018_WU183012.zip")
inputAnnotation <- file.path(datadir, "annotation_ComboCourse_p2691_March_2018_WU183012.xlsx")
startdata <- LFQService::tidyMQ_Peptides(inputMQfile)
startdata <- startdata %>% mutate(proteins = case_when(proteins == "" ~ leading.razor.protein, TRUE ~ proteins))

undebug(tidyMQ_top_protein_name)
startdata <- LFQService::tidyMQ_top_protein_name(startdata, prot_col = "proteins")

#startdata <- LFQService::tidyMQ_top_protein_name(startdata, prot_col = "leading.razor.protein")
#startdata <- startdata %>% mutate(top_protein = leading.razor.protein)


annotation <- readxl::read_xlsx(inputAnnotation)
annotation$experiment = 2691

```

Add annotation information data to configuration and data.

```{r}
config <- create_MQ_peptide_Configuration()
data <- inner_join(annotation, startdata)

config$table$factors[["condition_"]] = "condition"
config$table$factors[["experiment"]] = "experiment"
config$table$factors[["Run_ID"]] = "Run_ID"
config$table$factorDepth <- 1
```

Preprocess and normalize data.

```{r}
adata <- setup_analysis(data, config)
xx <- toWideConfig(adata, config)

adata <- filter_proteins_by_peptide_count(adata, config)$data
adata <- LFQService::remove_small_intensities(adata,config)


p2691 <- LFQData$new(adata, config)
p2691$get_Plotter()$intensity_distribution_density()

tr <- p2691$get_Transformer()
p2691 <- tr$log2_robscale()

pl <- p2691$get_Plotter()
pl$intensity_distribution_density()
#dev.off()
pha2691 <- pl$NA_heatmap()

```

```{r fig.cap="missing value heatmap for project 2691"}
print(pha2691)
```


# Prepare p2370 dataset


```{r}
datadir <- file.path(find.package("prolfquaData") , "quantdata")
inputMQfile <-  file.path(datadir, "MAXQuant_ComboCourse_p2370_March_2017_WU183008.zip")
inputAnnotation <- file.path(datadir, "annotation_ComboCourse_p2370_March_2017_WU183008.xlsx")

startdata <- LFQService::tidyMQ_Peptides(inputMQfile)
startdata <- startdata %>% mutate(proteins = case_when(proteins == "" ~ leading.razor.protein, TRUE ~ proteins))
startdata <- LFQService::tidyMQ_top_protein_name(startdata, prot_col = "proteins")

annotation <- readxl::read_xlsx(inputAnnotation)
annotation$experiment <- 2370
```

Code snipped below shows how to add annotation to the data and configuration.

```{r}
data <- inner_join(annotation, startdata)

config <- create_MQ_peptide_Configuration()

config$table$factors[["condition_"]] = "condition"
config$table$factors[["experiment"]] = "experiment"
config$table$factors[["Run_ID"]] = "Run_ID"
config$table$factorDepth <- 1
```

```{r}
adata <- setup_analysis(data, config)
xx <- toWideConfig(adata, config)
adata <- filter_proteins_by_peptide_count(adata, config)$data
adata <- LFQService::remove_small_intensities(adata,config)

p2370 <- LFQData$new(adata, config)
p2370$get_Plotter()$intensity_distribution_density()

tr <- p2370$get_Transformer()
p2370 <- tr$log2_robscale()
pl <- p2370$get_Plotter()
ha2370 <- pl$NA_heatmap()
```

```{r, fig.cap="Intensity distribution after normalization, p2370"}
pl$intensity_distribution_density()
```

```{r, fig.cap="Heatmap of missing values p2370."}
print(ha2370)
```


# Merge both dataset

To simulate a factorial design we merge both datasets.

```{r}

outpath <- "results_modelling"
data <- bind_rows(p2691$data, p2370$data)
sum(is.na(p2691$data$protein_Id))
sum(is.na(p2370$data$protein_Id))


# creates default configuration

config <- p2691$config$clone(deep=TRUE)
config$table$factorDepth <- 2
```

## QC for merged dataset 

```{r}
pl$intensity_distribution_density()
```


```{r}
pMerged <- LFQData$new(data, config)

bb <- pMerged$complete_cases()
sum(is.na(bb$protein_Id))
pl <- pMerged$get_Plotter()
pl$sample_correlation()
ph <- pl$heatmap_cor()
pha <- pl$NA_heatmap()
```

```{r fig.cap="sample correlation heatmap merged"}
print(ph)
```

```{r fig.cap ="Missing data heatmap for merged data"}
print(pha)
```

```{r}
su <- pMerged$get_Summariser()
su$hierarchy_counts()
```

```{r, fig.cap = "Number of Proteins or peptides per sample in merged dataset"}
su$hierarchy_counts_sample("plot")
```



```{r}
# specify model definition
modelName  <- "Model"
model <- "~ condition_*batch_ + (1|peptide_Id)  + (1|sampleName)"
DEBUG <- TRUE

Contrasts <- c("Glucose - Ethanol" = "condition_Glucose - condition_Ethanol",
               "p2370 - p2691" = "batch_p2370 - batch_p2691",
               "Glucose_vs_Ethanol_gv_p2370" = "`condition_Glucose:batch_p2370` - `condition_Ethanol:batch_p2370`",
               "Glucose_vs_Ethanol_gv_p2691" = "`condition_Glucose:batch_p2691` - `condition_Ethanol:batch_p2691`",
               "Interaction" = "`Glucose_vs_Ethanol_gv_p2370` - `Glucose_vs_Ethanol_gv_p2691`"
)




```
